% chapter4_system_design.tex
% BondHealth Healthcare Platform - System Design Chapter
% COMPLETELY SELF-CONTAINED - NO IMAGES NEEDED
% Follows formatting: 14pt headings, 12pt body, Times New Roman, 1.15 line spacing, justified

\chapter{System Design}

This chapter presents the comprehensive system design of the BondHealth healthcare platform, detailing the architectural decisions, data flow, database schema, API design, module specifications, security architecture, and third-party integrations that form the foundation of the unified digital healthcare ecosystem. The system design follows industry best practices for healthcare applications, ensuring scalability, security, maintainability, and seamless connectivity between patients, doctors, hospital administrators, and laboratory technicians.

% =============================================================================
% 4.1 SYSTEM ARCHITECTURE OVERVIEW
% =============================================================================

\section{System Architecture Overview}

The BondHealth platform is built on a robust three-tier architecture that ensures scalability, maintainability, and security while providing seamless connectivity between patients, doctors, hospital administrators, and laboratory technicians. The architecture follows industry best practices for healthcare applications, with clear separation of concerns between presentation, business logic, and data persistence layers. This three-tier approach enables independent development, deployment, and scaling of each layer while maintaining secure communication between them.

\subsection{Presentation Layer (Client Tier)}

The presentation layer serves as the user interface through which all stakeholders interact with the BondHealth platform. It utilizes server-side rendered HTML with Tailwind CSS framework to deliver responsive, mobile-first design across all devices including desktops, tablets, and smartphones. The server-side rendering approach offers several critical advantages for healthcare applications:

\begin{itemize}
    \item \textbf{Fast Initial Page Loads}: HTML is generated on the server and sent to the client, reducing client-side processing and improving time-to-interactive metrics, which is crucial for emergency healthcare scenarios where every second counts.
    
    \item \textbf{Improved SEO Capabilities}: Search engines can easily crawl and index server-rendered content, making healthcare services more discoverable to patients searching for medical assistance in their area.
    
    \item \textbf{Enhanced Security}: Business logic remains on the server, minimizing client-side exposure of sensitive healthcare algorithms and reducing attack vectors for malicious users attempting to access patient data.
    
    \item \textbf{Consistent User Experience}: The interface adapts seamlessly to different screen sizes and devices, ensuring healthcare professionals can access patient data from any location - whether at their desk, on hospital rounds with tablets, or responding to emergencies via mobile phones.
    
    \item \textbf{Accessibility Compliance}: The presentation layer implements WCAG 2.1 guidelines, ensuring the platform is usable by individuals with disabilities, including screen reader compatibility and keyboard navigation support for users with mobility impairments.
\end{itemize}

The user interface is organized into role-specific dashboards, each optimized for the unique workflows of different user types. Patient dashboards focus on appointment management and medical record access, doctor dashboards prioritize clinical workflow efficiency, admin dashboards provide comprehensive hospital management capabilities, and lab technician dashboards streamline test processing and result reporting. Each dashboard presents information in a clear, hierarchical manner with the most critical information displayed prominently.

\subsection{Application Layer (Business Tier)}

The application layer forms the core of the BondHealth platform, implementing all business logic, validation rules, workflow orchestration, and security controls. It is built using Node.js with the Express framework, providing a lightweight, event-driven architecture ideal for handling concurrent healthcare transactions. This layer processes all incoming requests, applies business rules, and coordinates responses between the presentation and data layers.

\textbf{Key Components of the Application Layer:}

\begin{enumerate}
    \item \textbf{RESTful API Endpoints}: Well-defined endpoints following REST conventions handle all client requests, with appropriate HTTP methods (GET, POST, PUT, DELETE) and status codes ensuring predictable interactions. Each endpoint is documented using OpenAPI/Swagger specifications.
    
    \item \textbf{Authentication Middleware}: JWT-based authentication verifies user identity for every protected request, with middleware functions that extract tokens from HTTP-only cookies and validate them against the active sessions store. Failed authentication attempts are logged and can trigger account lockout mechanisms.
    
    \item \textbf{Role-Based Access Control (RBAC)}: Middleware functions check user roles against required permissions for each endpoint, ensuring patients cannot access doctor functions, administrators can manage hospital operations, and laboratory technicians can only access test-related features. This implements the principle of least privilege throughout the application.
    
    \item \textbf{Request Validation Layer}: All incoming data is validated against predefined schemas using express-validator, preventing malformed or malicious data from reaching the database. Validation includes type checking, format validation (email, phone, dates), sanitization to remove potentially harmful content, and business rule enforcement such as checking appointment availability before booking.
    
    \item \textbf{Error Handling Middleware}: Centralized error handling captures, logs, and appropriately responds to exceptions, providing meaningful error messages to clients while masking sensitive implementation details. Different error types (validation errors, authentication failures, server errors) receive appropriate HTTP status codes and user-friendly messages.
    
    \item \textbf{Microservices-Oriented Design}: The application is organized into loosely coupled modules that can be independently scaled based on demand. The appointment booking service, payment processing service, real-time chat service, and notification service can each be deployed and scaled separately, allowing efficient resource utilization during peak usage periods such as morning appointment booking rushes.
    
    \item \textbf{Request Rate Limiting}: Protection against denial-of-service attacks through configurable rate limiting per IP address and per user account, with stricter limits on authentication endpoints to prevent brute-force attacks. Different endpoints have different rate limits based on their sensitivity and resource requirements.
\end{enumerate}

The application layer also implements comprehensive logging and monitoring, capturing request/response cycles, performance metrics, and error conditions for operational visibility and debugging. All logs are structured and centralized for easy analysis and alerting.

\subsection{Data Layer (Database Tier)}

The data layer manages all persistent storage for the BondHealth platform, implementing robust data modeling, access patterns, and security controls. Neon PostgreSQL serves as the primary database, leveraging serverless architecture for automatic scaling based on demand, eliminating the need for capacity planning while ensuring consistent performance during traffic spikes.

\textbf{Database Technology Stack Features:}

\begin{itemize}
    \item \textbf{Neon PostgreSQL}: Serverless PostgreSQL offering with built-in connection pooling, automatic failover, and point-in-time recovery capabilities. The serverless architecture means the database scales to zero during low activity periods (such as nights) and instantly scales up during peak loads, optimizing costs while maintaining performance. This is particularly valuable for healthcare applications that may have unpredictable usage patterns.
    
    \item \textbf{Connection Pooling}: Implemented through the \texttt{@neondatabase/serverless} package, maintaining a configurable pool of database connections that are reused across requests. This eliminates the overhead of establishing new connections for each operation and prevents connection exhaustion under load. The pool size is dynamically adjusted based on current demand.
    
    \item \textbf{UUID Primary Keys}: All tables use UUID v4 as primary keys instead of sequential integers, providing several advantages for healthcare applications:
    \begin{itemize}
        \item Distributed uniqueness allows safe data merging across multiple hospital instances if the platform expands
        \item Prevents enumeration attacks where sequential IDs would reveal data volume and patterns
        \item Enables offline record creation with guaranteed uniqueness when synchronized later
        \item Simplifies sharding and distributed database architectures
    \end{itemize}
    
    \item \textbf{PostgreSQL Array Data Types}: Multi-value fields such as medical conditions, allergies, and doctor availability days are stored as native PostgreSQL arrays, providing efficient storage and querying capabilities while maintaining data integrity through type enforcement. This eliminates the need for separate junction tables for simple lists.
    
    \item \textbf{Foreign Key Constraints}: Referential integrity is enforced through foreign key relationships between tables, ensuring that orphaned records cannot exist and that operations like cascading deletes maintain data consistency. All relationships are explicitly defined with appropriate cascade rules.
    
    \item \textbf{SSL Encryption}: All database connections require SSL/TLS encryption, ensuring that sensitive patient data transmitted between the application and database servers cannot be intercepted or read in transit. Certificate validation ensures connection authenticity.
    
    \item \textbf{Automated Backups}: Neon PostgreSQL provides continuous backup capabilities with point-in-time recovery, enabling restoration to any moment within the retention period - a critical requirement for healthcare data compliance and disaster recovery planning.
\end{itemize}

The database schema is normalized to reduce redundancy while being optimized for healthcare-specific query patterns through strategic denormalization where performance demands it. Indexes are carefully designed to support common query patterns without excessive overhead.

% =============================================================================
% 4.2 DATA FLOW DIAGRAMS
% =============================================================================

\section{Data Flow Diagrams}

Data Flow Diagrams (DFDs) provide a graphical representation of how information moves through the BondHealth platform, illustrating the relationships between external entities, processes, and data stores. This section describes the three most critical workflows in the system: appointment booking, lab report processing, and payment processing. Each description follows the Gane-Sarson notation with external entities, processes, data stores, and data flows clearly defined.

\subsection{Appointment Booking Data Flow}

The appointment booking process represents one of the most frequently used workflows in the BondHealth platform, enabling patients to schedule consultations with doctors based on real-time availability. The complete data flow encompasses authentication, doctor search, availability checking, appointment creation, notification delivery, and dashboard updates across multiple system components.

\textbf{Detailed Process Description:}

\begin{enumerate}
    \item \textbf{Patient Authentication}: The patient provides login credentials through the login form. The system validates these credentials against the Users data store, comparing the provided password with the stored bcrypt hash. Upon successful authentication, the system generates a JWT token containing user identity and role information, stores it in an HTTP-only cookie, and establishes an authenticated session in the server-side session store. Failed attempts are logged for security monitoring, and multiple failures trigger temporary account lockout.
    
    \item \textbf{Doctor Search and Filtering}: The authenticated patient accesses the doctor search interface and applies filters such as specialization, hospital location, availability date, gender preference, languages spoken, and patient ratings. The system queries the Doctors data store, applying the specified filters and returning matching doctor profiles including their consultation fees, qualifications, experience, and available time slots. Results are paginated for performance and sorted by relevance or patient preference.
    
    \item \textbf{Real-time Availability Check}: When the patient selects a specific doctor and date, the system queries the Appointments data store to retrieve all existing appointments for that doctor on the selected date. It also checks the Doctor Leave data store to verify the doctor is not on leave during the requested period. The system calculates available time slots by subtracting booked appointments and leave periods from the doctor's configured working hours (stored as \texttt{available\_days} and \texttt{available\_time} arrays). This calculation respects appointment duration and buffer times between appointments.
    
    \item \textbf{Slot Selection and Booking}: The patient selects an available time slot and provides the reason for visit. The system validates that the slot remains available (preventing double-booking through database transactions with row-level locking), creates a new appointment record in the Appointments data store with status set to 'pending', and generates a unique appointment UUID and token number. The system also records the timestamp and IP address for audit purposes.
    
    \item \textbf{Payment Processing (if applicable)}: For paid consultations, the system initiates the payment workflow, creating a payment record in the Payments data store with status 'pending' and redirecting to the payment gateway. Upon successful payment confirmation via webhook, the appointment status is updated to 'confirmed'. If payment fails, the appointment status remains 'pending' and the patient is notified to retry.
    
    \item \textbf{Notification Generation}: The system triggers the Notification Module to send confirmation messages through multiple channels:
    \begin{itemize}
        \item Email confirmation with appointment details, doctor information, and calendar attachment (.ics file) for easy import
        \item SMS reminder with date, time, and location (or video link for telehealth)
        \item In-app notification visible on the patient dashboard with action buttons
        \item Push notification if the patient has enabled mobile alerts on their device
    \end{itemize}
    
    \item \textbf{Dashboard Updates}: The patient's dashboard is updated in real-time to display the new appointment in the "Upcoming Appointments" section. The doctor's dashboard is updated via WebSocket to show the new booking in their schedule with appropriate status indicators. Hospital administrators can view the appointment in their management interface for operational oversight.
\end{enumerate}

\textbf{Data Stores Involved:}
\begin{itemize}
    \item \textbf{Users}: Stores authentication credentials, user profiles, and role information for all system users
    \item \textbf{Patients}: Contains patient-specific data including medical conditions, allergies, and emergency contacts
    \item \textbf{Doctors}: Stores doctor profiles, specializations, consultation fees, and availability patterns
    \item \textbf{Appointments}: Maintains all appointment records with status, dates, times, and reasons for visit
    \item \textbf{Doctor Leave}: Tracks doctor leave periods that affect appointment availability calculations
    \item \textbf{Payments}: Records payment transactions linked to appointments for financial tracking
\end{itemize}

\subsection{Lab Report Processing Data Flow}

Laboratory technicians use this workflow to process patient test requests, enter results, upload supporting documents, and share reports with doctors and patients according to configured privacy preferences. The workflow ensures that test results are accurately recorded, properly prioritized, and securely delivered to authorized recipients while maintaining a complete audit trail.

\textbf{Detailed Process Description:}

\begin{enumerate}
    \item \textbf{Technician Authentication}: Lab technicians log into the system using their credentials. The authentication middleware verifies the user's role as 'lab' and grants access to lab-specific features. Failed authentication attempts are logged, and account lockouts are triggered after multiple failures. Successful logins record the technician's identity for audit purposes.
    
    \item \textbf{Patient Selection}: The technician accesses the patient search interface, which provides multiple lookup methods:
    \begin{itemize}
        \item Search by patient UUID or medical record number for precise matching
        \item Search by name with fuzzy matching to handle spelling variations
        \item Browse recent patients from the last 7 days for quick access to repeat patients
        \item Scan QR code from patient identification band (if using mobile/tablet interface) for error-free identification
    \end{itemize}
    The system queries the Patients data store and returns matching patient records with basic demographic information for confirmation.
    
    \item \textbf{Test Information Entry}: The technician selects the test type from a predefined list of laboratory tests, each with associated normal ranges, units, and required input fields. The system displays appropriate input fields based on test type:
    \begin{itemize}
        \item Numeric fields with unit selection for quantitative tests (e.g., blood glucose in mg/dL, cholesterol in mmol/L)
        \item Text areas for qualitative observations (e.g., urinalysis description, culture results)
        \item Multiple choice for categorical results (e.g., positive/negative, reactive/non-reactive, present/absent)
        \item Date/time fields for timed measurements (e.g., peak drug levels)
    \end{itemize}
    
    \item \textbf{Result Entry and Validation}: The technician enters test results with real-time validation:
    \begin{itemize}
        \item Range checking against normal values with visual indicators for out-of-range results (highlighted in yellow for abnormal, red for critical)
        \item Unit consistency verification to prevent unit conversion errors
        \item Required field completion validation before submission
        \item Critical value alerts that require confirmation before submission, with mandatory override reason
    \end{itemize}
    
    \item \textbf{Priority Assignment}: The technician sets the report priority level, which determines notification urgency and dashboard highlighting:
    \begin{itemize}
        \item \textbf{Normal}: Routine results, standard notification via email and in-app (blue indicator)
        \item \textbf{Urgent}: Results requiring prompt attention, email + SMS notification with escalation if not viewed (orange indicator)
        \item \textbf{Critical}: Life-threatening values requiring immediate action, SMS + phone call escalation protocol (red indicator with confirmation required and escalation to on-call staff)
    \end{itemize}
    
    \item \textbf{File Upload}: Supporting documents can be attached to the report:
    \begin{itemize}
        \item PDF reports from laboratory equipment with automatic metadata extraction
        \item Images of slides, cultures, or test strips for visual reference
        \item Scanned documents from external labs
    \end{itemize}
    The system validates file types (allowed: PDF, JPG, PNG, TIFF), enforces size limits (max 25MB per file), and performs virus scanning before storage. Files are uploaded to AWS S3/Azure Blob storage with encrypted filenames and access controls. Each file receives a unique identifier and is stored with checksums for integrity verification.
    
    \item \textbf{Sharing Preferences Configuration}: The technician specifies who can access the report:
    \begin{itemize}
        \item \textbf{Doctor Only}: Report visible only to the requesting doctor (default for most diagnostic tests)
        \item \textbf{Patient Only}: Report visible only to the patient (rare, used for self-ordered tests or direct-to-consumer results)
        \item \textbf{Both}: Report visible to both doctor and patient (common for routine results like blood counts)
        \item \textbf{Referring Physician}: Option to share with additional doctors if specified in the test order
    \end{itemize}
    These preferences are stored in the \texttt{shared\_with} field and enforced by the access control middleware at both the API and data levels.
    
    \item \textbf{Report Submission and Storage}: Upon submission, the system:
    \begin{itemize}
        \item Creates a new record in the Lab Reports data store with all entered data, including metadata
        \item Generates a unique report UUID for external reference and patient communication
        \item Updates the report status to 'completed' (or 'pending' if results are partial with follow-up tests)
        \item Associates the report with the patient, doctor, and lab technician records for complete traceability
        \item Triggers the notification workflow based on priority level and sharing preferences
        \item Records the submission timestamp and technician identity in the audit log
    \end{itemize}
    
    \item \textbf{Notification Delivery}: The Notification Module sends alerts based on priority and sharing preferences:
    \begin{itemize}
        \item Doctor receives in-app notification and email with direct link to the report
        \item Patient receives notification only if \texttt{shared\_with} includes 'patient' or 'both', with appropriate privacy messaging
        \item Critical results trigger immediate SMS and escalate to on-call protocols with confirmation tracking
        \item Report ready notifications include direct links to view/download the report with expiry times for security
    \end{itemize}
\end{enumerate}

\subsection{Payment Processing Data Flow}

The payment processing workflow handles all financial transactions within the BondHealth platform, including appointment fees, consultation charges, and potential refunds. The workflow integrates with external payment gateways while maintaining comprehensive internal records for reconciliation, audit purposes, and financial reporting.

\textbf{Detailed Process Description:}

\begin{enumerate}
    \item \textbf{Payment Initiation}: The payment workflow is triggered in one of several scenarios:
    \begin{itemize}
        \item Patient books an appointment requiring upfront payment (common for private consultations)
        \item Patient completes a consultation and is billed afterward (post-paid model)
        \item Patient pays an outstanding invoice from their account (settling balances)
        \item Admin processes a refund for a cancelled appointment (financial adjustment)
    \end{itemize}
    The system creates a payment record in the Payments data store with initial status 'pending' and generates a unique payment UUID for tracking.
    
    \item \textbf{Amount Calculation}: The system calculates the total amount due:
    \begin{itemize}
        \item Base consultation fee from the doctor's profile (may vary by appointment type)
        \item Applicable taxes (GST/VAT configured by hospital location)
        \item Any discounts or insurance adjustments (promotional codes, insurance coverage)
        \item Previous outstanding balances if applicable (consolidated billing)
    \end{itemize}
    The detailed breakdown is displayed to the patient for confirmation before proceeding.
    
    \item \textbf{Payment Method Selection}: The patient selects from available payment methods:
    \begin{itemize}
        \item \textbf{Credit/Debit Card}: Integrated with Stripe/Razorpay for secure card processing using tokenization
        \item \textbf{Insurance Billing}: Captures insurance information for later processing and claim submission
        \item \textbf{Cash}: Records cash payment with receipt generation (primarily for in-person visits)
        \item \textbf{Digital Wallets}: Support for Apple Pay, Google Pay, and local wallet solutions
        \item \textbf{UPI/Bank Transfer}: Region-specific payment methods (India, Southeast Asia)
    \end{itemize}
    
    \item \textbf{Payment Gateway Integration}: For card and digital wallet payments, the system:
    \begin{itemize}
        \item Creates a payment intent with the gateway using their API, specifying amount, currency, and metadata
        \item Receives a client secret for front-end tokenization (PCI compliance maintained by never handling raw card details)
        \item Implements 3D Secure authentication if required by the card issuer for additional security
        \item Handles asynchronous responses through webhooks for reliable processing
    \end{itemize}
    
    \item \textbf{Transaction Processing}: The gateway processes the payment and returns:
    \begin{itemize}
        \item Success response with transaction ID, authorization code, and timestamp
        \item Failure response with error codes and reasons (insufficient funds, card declined, etc.)
        \item Pending status for transactions under review (fraud checks, manual review)
    \end{itemize}
    The system updates the payment record with the transaction ID and current status, and logs the outcome.
    
    \item \textbf{Post-Payment Processing}: Upon successful payment:
    \begin{itemize}
        \item Appointment status is updated to 'confirmed' (if payment was for booking)
        \item Invoice is generated as PDF with all required details (hospital info, patient details, itemized charges, taxes)
        \item Invoice is stored in cloud storage and URL saved to payment record
        \item Receipt is emailed to the patient with the invoice attached
        \item Payment confirmation is displayed to the patient with transaction reference
        \item Accounting records are updated for hospital financial systems (automated journal entries)
    \end{itemize}
    
    \item \textbf{Refund Processing}: When cancellations occur, administrators can process refunds:
    \begin{itemize}
        \item Admin initiates refund through interface with reason selection (patient cancellation, duplicate payment, etc.)
        \item System validates refund eligibility based on cancellation policy and time elapsed
        \item Refund request is sent to payment gateway via API with appropriate parameters
        \item Gateway processes refund and returns confirmation with refund transaction ID
        \item Payment record is updated with refund status, reference, and timestamp
        \item Patient receives refund confirmation via email with credit note if applicable
    \end{itemize}
    
    \item \textbf{Webhook Handling}: The system implements webhook endpoints for asynchronous gateway notifications:
    \begin{itemize}
        \item Payment success/failure updates for transactions that were pending
        \item Dispute notifications when customers challenge charges
        \item Refund confirmations when gateway processes refunds
        \item Settlement reports for daily reconciliation
    \end{itemize}
    Webhooks update internal payment records and trigger appropriate business processes (e.g., marking appointments confirmed).
\end{enumerate}

% =============================================================================
% 4.3 USE CASE DIAGRAMS
% =============================================================================

\section{Use Case Diagrams}

Use case diagrams capture the functional requirements of the BondHealth platform from the perspective of different user roles, illustrating the interactions between actors and the system. Each use case represents a discrete unit of functionality that provides value to the user. This section describes all use cases for the four primary actor types: Patients, Doctors, Hospital Administrators, and Laboratory Technicians.

\subsection{Patient Use Cases}

Patients are the primary beneficiaries of the BondHealth platform, accessing a comprehensive suite of features designed to manage their healthcare journey from registration through treatment and follow-up. The patient use cases encompass all interactions that patients can perform within the system, organized into functional categories for clarity.

\textbf{Complete List of Patient Use Cases:}

\begin{enumerate}
    \item \textbf{Account Registration and Management}:
    \begin{itemize}
        \item Register new account with email verification and phone confirmation
        \item Login to system with username/password or biometric options
        \item Logout from system with session termination
        \item Reset forgotten password via email with secure tokens
        \item Update profile information including contact details and preferences
        \item Manage emergency contacts with relationship and priority
        \item Configure notification preferences per channel (email, SMS, push)
        \item Delete account with data export option (GDPR compliance)
    \end{itemize}
    
    \item \textbf{Profile and Health Information}:
    \begin{itemize}
        \item View personal profile summary with quick access to key information
        \item Update demographic information (address, phone, insurance details)
        \item Add/edit medical conditions with ICD-10 coding for accuracy
        \item Add/edit allergies with severity levels and reaction descriptions
        \item Upload insurance information including policy numbers and coverage
        \item View medical history timeline with chronological organization
        \item Download complete medical record as PDF for sharing with other providers
    \end{itemize}
    
    \item \textbf{Doctor Discovery and Selection}:
    \begin{itemize}
        \item Browse doctors by specialization with filtering options
        \item Search doctors by name, hospital, or location with autocomplete
        \item Filter doctors by availability, ratings, gender, languages, and accepted insurance
        \item View doctor profiles with qualifications, experience, patient reviews, and fees
        \item Read patient reviews and ratings with verification badges
        \item Check doctor availability calendar with real-time slot visibility
        \item Save favorite doctors for quick access to future appointments
    \end{itemize}
    
    \item \textbf{Appointment Management}:
    \begin{itemize}
        \item Book new appointment with selected doctor and preferred time
        \item Select preferred date and time from available slots with visual indicators
        \item Specify reason for visit with structured templates (new, follow-up, urgent)
        \item View all upcoming appointments in chronological order with countdown
        \item View past appointment history with visit summaries and notes
        \item Reschedule existing appointment with reason tracking and availability check
        \item Cancel appointment with reason and confirmation
        \item Add appointment to personal calendar (Google, Apple, Outlook) with one click
        \item Set appointment reminders with configurable timing
    \end{itemize}
    
    \item \textbf{Medical Records Access}:
    \begin{itemize}
        \item View all lab reports with test details and reference ranges
        \item Filter reports by date range, test type, ordering doctor
        \item Download report PDFs for offline access or sharing
        \item Share reports with other healthcare providers via secure link
        \item View report trends over time with graphical visualization (for recurring tests)
        \item Access radiology images and reports with zoom and measurement tools
        \item View vaccination records with due dates for boosters
    \end{itemize}
    
    \item \textbf{Prescription Management}:
    \begin{itemize}
        \item View active prescriptions with medicine details and dosage instructions
        \item View prescription history including expired and completed medications
        \item Request prescription refill with automatic doctor notification
        \item Download prescription PDF for pharmacy use
        \item Set medicine reminders with dosage schedules and adherence tracking
        \item View medication history with start/end dates and prescriber information
        \item Receive refill reminders before medication runs out (configurable days before)
    \end{itemize}
    
    \item \textbf{Payment and Billing}:
    \begin{itemize}
        \item Make payment for appointments using preferred payment method
        \item View payment history with transaction details and status
        \item Download invoices and receipts for reimbursement or records
        \item Save payment methods for future use with tokenization
        \item Request refunds for cancelled appointments with automatic eligibility check
        \item View outstanding balances and due dates
        \item Set up auto-pay for recurring appointments or subscriptions
    \end{itemize}
    
    \item \textbf{Communication}:
    \begin{itemize}
        \item Chat with doctors securely with message history and attachment support
        \item Receive notifications for all relevant events (appointment reminders, report ready)
        \item View message history with search and filtering capabilities
        \item Send files and images in chat (symptoms photos, previous reports)
        \item Receive broadcast messages from hospital (announcements, health tips)
        \item Provide feedback and ratings after consultations with detailed comments
    \end{itemize}
    
    \item \textbf{Health Management}:
    \begin{itemize}
        \item Set health reminders (medication, checkups, screenings)
        \item Track vital signs if integrated with wearables (optional feature)
        \item Complete pre-visit questionnaires before appointments
        \item View health tips and educational content personalized to conditions
        \item Manage family member accounts for dependents (children, elderly parents)
    \end{itemize}
\end{enumerate}

\subsection{Doctor Use Cases}

Doctors utilize the BondHealth platform to manage their clinical workflow efficiently, accessing patient information, recording observations, prescribing medications, and communicating with patients. The doctor use cases are designed to minimize administrative overhead while maximizing time available for direct patient care.

\textbf{Complete List of Doctor Use Cases:}

\begin{enumerate}
    \item \textbf{Account and Profile Management}:
    \begin{itemize}
        \item Login to system with secure credentials and optional 2FA
        \item Update professional profile including bio and clinical interests
        \item Modify consultation fees for different appointment types
        \item Update availability schedule (working days, hours, and break times)
        \item Upload/change profile photo for patient recognition
        \item Add/update qualifications and certifications with verification
        \item Set consultation modes (in-person, video, both) with availability per mode
        \item Configure notification preferences for different event types
    \end{itemize}
    
    \item \textbf{Dashboard Overview}:
    \begin{itemize}
        \item View personalized dashboard with summary cards and key metrics
        \item See today's appointment count and schedule with time blocks
        \item View pending lab reports count with priority breakdown
        \item Check unread messages count with previews
        \item Monitor upcoming leave status and approval state
        \item View patient statistics (total patients, new patients this week)
        \item Access quick actions for common tasks (prescribe, view reports)
    \end{itemize}
    
    \item \textbf{Appointment Management}:
    \begin{itemize}
        \item View today's complete schedule with detailed time slots and patient info
        \item View upcoming appointments for future dates with filtering
        \item View appointment history with past patients and visit reasons
        \item Start consultation for each appointment, recording start time
        \item Mark patient as arrived/checked-in for in-person visits
        \item Complete consultation and close appointment with summary notes
        \item View patient wait times and queue status for clinic flow management
        \item Block time slots for personal reasons or administrative tasks
    \end{itemize}
    
    \item \textbf{Patient Management}:
    \begin{itemize}
        \item View complete patient list with pagination and sorting options
        \item Search patients by name, ID, or contact with fuzzy matching
        \item Filter patients by recent visits, conditions, or demographics
        \item Access patient medical history with timeline visualization
        \item View past consultations with complete notes and diagnoses
        \item See patient demographics and emergency contacts for reference
        \item Add patients to favorites for quick access to frequent visitors
        \item View patient visit timeline with clinical events in order
    \end{itemize}
    
    \item \textbf{Lab Report Management}:
    \begin{itemize}
        \item View pending lab reports queue with count and priority indicators
        \item Filter reports by priority (normal/urgent/critical) and date
        \item View complete report details with results, reference ranges, and flags
        \item Add clinical findings and interpretations to reports
        \item Correlate results with patient history and previous values
        \item Flag abnormal results for follow-up or discussion
        \item Request additional tests if needed based on findings
        \item Mark reports as reviewed with timestamp for audit
        \item Download or print reports for physical records
    \end{itemize}
    
    \item \textbf{Prescription Management}:
    \begin{itemize}
        \item Create new prescriptions with medicine selection from hospital formulary
        \item Select medicines with dosage suggestions and common prescriptions
        \item Specify dosage, frequency, and duration with standardized options
        \item Add special instructions for patient (e.g., ``take with food'')
        \item Set prescription validity period and refill limits
        \item Renew existing prescriptions with one-click and date adjustment
        \item View patient's active medications to avoid conflicts
        \item Check for drug interactions automatically (integrated database)
        \item Print or email prescription to patient with pharmacy options
        \item Add digital signature for authenticity
    \end{itemize}
    
    \item \textbf{Leave Management}:
    \begin{itemize}
        \item Apply for leave with date range and reason selection
        \item Specify reason for leave (vacation, conference, sick)
        \item View leave balance if tracked by hospital policy
        \item Check leave approval status in real-time
        \item Cancel pending leave requests if plans change
        \item View upcoming approved leaves on calendar
        \item Set auto-reply for leave periods with alternative contact
    \end{itemize}
    
    \item \textbf{Communication}:
    \begin{itemize}
        \item Chat with patients securely with end-to-end encryption
        \item View message history with each patient organized by date
        \item Send file attachments (images, PDFs) for reference
        \item Receive notifications for new messages with preview
        \item Send broadcast messages to multiple patients (announcements)
        \item Conduct video consultations if integrated (telehealth)
        \item Share educational materials with patients (articles, videos)
    \end{itemize}
    
    \item \textbf{Clinical Documentation}:
    \begin{itemize}
        \item Write consultation notes with templates for efficiency
        \item Create diagnosis records with ICD-10 codes for billing
        \item Document treatment plans with follow-up schedule
        \item Record vital signs measurements during consultations
        \item Generate referral letters to specialists with patient context
        \item Complete medical certificates with digital signature
        \item Maintain procedure notes for any in-office procedures
    \end{itemize}
\end{enumerate}

\subsection{Admin Use Cases}

Hospital administrators oversee the operational aspects of the BondHealth platform, managing users, inventory, and system configuration. The admin use cases provide comprehensive tools for hospital management, ensuring smooth operation of all healthcare services and enabling data-driven decision making.

\textbf{Complete List of Admin Use Cases:}

\begin{enumerate}
    \item \textbf{Hospital Management}:
    \begin{itemize}
        \item Register new hospital/profile with all required details
        \item Update hospital information (address, contact, facilities)
        \item Manage hospital departments with heads and staff
        \item Configure hospital working hours and holiday schedule
        \item Set hospital policies (cancellation, refund, visit duration)
        \item Manage hospital locations/branches with individual settings
        \item Upload hospital logos and branding for patient-facing materials
        \item Configure tax rates and currency settings per location
    \end{itemize}
    
    \item \textbf{Doctor Management}:
    \begin{itemize}
        \item Add new doctors to the system with complete profile
        \item Verify doctor credentials and licenses with document upload
        \item Assign doctors to departments and specializations
        \item Set doctor consultation fees for different appointment types
        \item Configure doctor schedules with recurring patterns
        \item Edit doctor profiles as needed for updates
        \item Deactivate/reactivate doctor accounts for leaves or separation
        \item View doctor performance metrics (patients seen, ratings, revenue)
        \item Manage doctor leave approvals with calendar view
        \item Handle doctor resignation/termination with patient reassignment
    \end{itemize}
    
    \item \textbf{Patient Management}:
    \begin{itemize}
        \item View all patient registrations with demographic filters
        \item Search patients by various criteria (name, ID, phone, email)
        \item Assist with patient account issues (password reset, unlocking)
        \item Merge duplicate patient records with data preservation
        \item Manage patient data access requests for compliance
        \item Handle patient complaints with tracking and resolution
        \item Export patient data for reporting or migration
    \end{itemize}
    
    \item \textbf{Lab Technician Management}:
    \begin{itemize}
        \item Add new lab technicians with credentials and qualifications
        \item Assign technicians to labs/departments based on expertise
        \item Manage technician roles and permissions within lab module
        \item View technician workload and performance metrics
        \item Deactivate technician accounts as needed
        \item Manage technician schedules and shift assignments
    \end{itemize}
    
    \item \textbf{Department and Specialization Management}:
    \begin{itemize}
        \item Create new medical departments with descriptions
        \item Add specializations under departments with standard codes
        \item Assign doctors to specializations with primary/secondary
        \item Configure department-specific settings (consultation duration, fees)
        \item Manage department heads with reporting relationships
        \item View department performance metrics (volume, revenue, satisfaction)
    \end{itemize}
    
    \item \textbf{Medicine Inventory Management}:
    \begin{itemize}
        \item Add new medicines to inventory with complete details
        \item Update medicine quantities with adjustment reasons
        \item Set reorder levels and automatic alerts
        \item Track expiry dates with notification before expiry
        \item Manage medicine pricing with retail and insurance rates
        \item View inventory reports with usage patterns and forecasts
        \item Process medicine requisitions from departments
        \item Manage suppliers and purchase orders with tracking
        \item Track medicine usage patterns for procurement planning
    \end{itemize}
    
    \item \textbf{Appointment Oversight}:
    \begin{itemize}
        \item View all appointments across hospital with comprehensive filters
        \item Filter appointments by status, doctor, date, department
        \item Handle appointment conflicts with resolution tools
        \item View no-show reports with patient history
        \item Monitor appointment utilization (fill rates, cancellations)
        \item Generate appointment analytics with visual dashboards
        \item Manage waitlist for popular slots with auto-assignment
    \end{itemize}
    
    \item \textbf{Leave Management}:
    \begin{itemize}
        \item View all leave requests with status and details
        \item Filter by status (pending, approved, rejected) for workflow
        \item Approve or reject leave requests with comment capability
        \item View leave calendar showing coverage across departments
        \item Manage leave policies and balances per employee
        \item Generate leave reports for HR and planning
        \item Handle emergency leave situations with override
    \end{itemize}
    
    \item \textbf{Financial Management}:
    \begin{itemize}
        \item View revenue reports with daily, weekly, monthly summaries
        \item Monitor payment collections by method and status
        \item Process refunds with approval workflow
        \item View outstanding payments and aging reports
        \item Generate financial statements (P\&L, balance sheet)
        \item Manage insurance claims with tracking and follow-up
        \item Configure payment gateway settings and fee structures
        \item Reconcile transactions with bank statements
    \end{itemize}
    
    \item \textbf{Reporting and Analytics}:
    \begin{itemize}
        \item Generate operational reports (appointments, utilization, wait times)
        \item View dashboard statistics with real-time updates
        \item Export data for external analysis in various formats
        \item Monitor key performance indicators with trend charts
        \item Track patient satisfaction metrics from feedback
        \item Analyze appointment trends for capacity planning
        \item View doctor performance analytics with comparisons
        \item Generate compliance reports for regulatory requirements
    \end{itemize}
    
    \item \textbf{System Configuration}:
    \begin{itemize}
        \item Manage user roles and permissions with granular controls
        \item Configure notification templates for all channels
        \item Set system parameters (timeouts, limits, defaults)
        \item Manage audit logs with search and export
        \item Configure integration settings for third-party services
        \item Perform system backups and test restoration
        \item Monitor system health and performance metrics
    \end{itemize}
\end{enumerate}

\subsection{Lab Technician Use Cases}

Laboratory technicians have specialized use cases focused on test processing, result entry, and report management. These use cases are designed to streamline laboratory workflows while maintaining accuracy, quality control, and efficient communication with requesting doctors.

\textbf{Complete List of Lab Technician Use Cases:}

\begin{enumerate}
    \item \textbf{Account Management}:
    \begin{itemize}
        \item Login to system with secure credentials and role verification
        \item Update personal profile (contact details, preferences)
        \item Change password with security requirements
        \item View dashboard with work summary and daily statistics
        \item Logout from system with session cleanup
    \end{itemize}
    
    \item \textbf{Dashboard Overview}:
    \begin{itemize}
        \item View today's test statistics (total, completed, pending)
        \item See pending reports count with priority breakdown
        \item View completed reports count for the day
        \item Monitor urgent/critical cases requiring immediate attention
        \item Check workload distribution across technicians
        \item View recent activity with timestamps
        \item Access quick actions for common tasks (new report, search)
    \end{itemize}
    
    \item \textbf{Patient Management}:
    \begin{itemize}
        \item Search for patients by ID, name, or contact with multiple criteria
        \item View patient demographics for identification
        \item Access patient test history for context
        \item View previous test results for comparison
        \item Verify patient identity before testing (name, DOB, ID)
        \item Add new patient if authorized (emergency or walk-in)
        \item Update patient contact information if provided
    \end{itemize}
    
    \item \textbf{Lab Report Creation}:
    \begin{itemize}
        \item Create new lab report with automatic ID generation
        \item Select patient from search results with confirmation
        \item Choose test type from standardized list with categories
        \item Enter test date and time (defaults to now)
        \item Add ordering/referring doctor from search
        \item Enter test results with appropriate input fields
        \item Add qualitative observations and interpretations
        \item Upload supporting files (PDFs, images) with validation
        \item Set report priority (normal/urgent/critical) with impact explanation
        \item Configure sharing preferences (doctor/patient/both)
        \item Add comments or notes for internal use
        \item Save draft for later completion if interrupted
    \end{itemize}
    
    \item \textbf{Result Entry and Validation}:
    \begin{itemize}
        \item Enter numerical results with unit selection from predefined options
        \item Receive real-time validation against reference ranges
        \item Get alerts for out-of-range values with severity indication
        \item Confirm critical values before submission with override reason
        \item Enter qualitative results from standardized dropdowns
        \item Add free-text interpretations and clinical correlations
        \item Flag abnormal results for attention with color coding
        \item Correlate results with previous tests for trending
    \end{itemize}
    
    \item \textbf{File Management}:
    \begin{itemize}
        \item Upload report files (PDF, images) with drag-and-drop
        \item Preview uploaded files before finalizing
        \item Replace/update files if errors detected
        \item Delete incorrect attachments with confirmation
        \item View file upload history with versions
        \item Ensure file naming follows conventions automatically
        \item Verify file readability and quality indicators
    \end{itemize}
    
    \item \textbf{Report Management}:
    \begin{itemize}
        \item View all created reports with filters and sorting
        \item Filter reports by status (draft, pending, completed, reviewed)
        \item Search reports by patient, date, test type, priority
        \item Edit pending reports with version tracking
        \item Submit completed reports with final review
        \item View report details in comprehensive format
        \item Print reports with configured layouts
        \item Download reports as PDF for distribution
        \item Resend reports if delivery failed
        \item Track report delivery status (viewed by doctor/patient)
    \end{itemize}
    
    \item \textbf{Quality Control}:
    \begin{itemize}
        \item Verify results against quality standards and controls
        \item Document any quality issues or deviations
        \item Flag results requiring re-testing with reasons
        \item Participate in proficiency testing records
        \item Maintain equipment logs if integrated
        \item Record reagent lots and expiry dates for traceability
    \end{itemize}
    
    \item \textbf{Communication}:
    \begin{itemize}
        \item Receive notifications for new test requests
        \item Send alerts for critical results with confirmation
        \item Communicate with doctors about results or issues
        \item Receive feedback on report quality and timeliness
        \item Participate in lab team communications (announcements)
    \end{itemize}
    
    \item \textbf{Workflow Management}:
    \begin{itemize}
        \item Prioritize work based on urgency and received time
        \item Manage test queue with drag-and-drop prioritization
        \item Track turnaround times for performance monitoring
        \item Hand off work to next shift with notes
        \item Report equipment issues with ticketing
        \item Request supplies when low
    \end{itemize}
\end{enumerate}

% =============================================================================
% 4.4 DATABASE SCHEMA DESIGN
% =============================================================================

\section{Database Schema Design (Neon PostgreSQL)}

The BondHealth platform utilizes Neon PostgreSQL, a serverless PostgreSQL solution that provides automatic scaling, high availability, and connection pooling optimized for modern cloud applications. The database design implements healthcare industry best practices including UUID primary keys for distributed uniqueness, array data types for efficient multi-value storage, and comprehensive foreign key constraints for referential integrity. SSL connections are enforced for all database communications, ensuring data security in transit and compliance with healthcare data protection regulations such as HIPAA and GDPR.

\textbf{Database Design Principles:}

\begin{itemize}
    \item \textbf{Normalization}: The schema follows third normal form (3NF) to eliminate data redundancy while maintaining efficient query performance through strategic indexing. This balance ensures data integrity without sacrificing speed for common queries.
    
    \item \textbf{Data Integrity}: Foreign key constraints with appropriate cascade rules ensure that related data remains consistent across all tables. Cascading deletes are used judiciously where appropriate (e.g., removing patient data when user is deleted), while restrict rules prevent accidental data loss in other relationships.
    
    \item \textbf{Audit Capabilities}: \texttt{created\_at} and \texttt{updated\_at} timestamps on all tables provide audit trails for regulatory compliance and debugging. Additional audit tables track changes to sensitive data for forensic analysis.
    
    \item \textbf{Performance Optimization}: Strategic indexes on frequently queried columns ensure rapid data retrieval even as the database scales to millions of records. Composite indexes support complex queries with multiple conditions.
    
    \item \textbf{PostgreSQL-Specific Features}: Leveraging array data types for multi-value attributes, JSON fields where flexible schemas are beneficial, and full-text search capabilities for enhanced search functionality across patient records and doctor profiles.
\end{itemize}

The following tables form the core of the BondHealth database schema, each designed with specific healthcare requirements in mind.

\begin{table}[h]
    \centering
    \caption{Users Schema - Core table storing authentication credentials and base user information for all system users. This table serves as the foundation for role-based access control and contains the minimal data required for authentication and session management. Each user has a unique identifier and can be associated with a hospital if they are staff (doctors, admins, lab technicians).}
    \label{tab:users}
    \begin{tabular}{|p{3cm}|p{5cm}|p{3cm}|p{3cm}|}
        \hline
        \textbf{Property Name} & \textbf{Description} & \textbf{Data Type} & \textbf{Constraints} \\
        \hline
        user\_id & Unique identifier for each user, generated as UUID v4 to ensure global uniqueness across distributed systems & UUID & PRIMARY KEY, DEFAULT uuid\_generate\_v4() \\
        \hline
        username & Unique username chosen by user for login, must be between 3-50 characters, alphanumeric with underscores allowed & VARCHAR(100) & NOT NULL, UNIQUE \\
        \hline
        email & User's email address used for login, notifications, and password recovery. Verified during registration & VARCHAR(255) & NOT NULL, UNIQUE \\
        \hline
        password\_hash & Bcrypt hash of user's password with 10 salt rounds, never stores plaintext passwords & VARCHAR(255) & NOT NULL \\
        \hline
        role & User's role in the system determining access permissions. Values: 'patient', 'doctor', 'admin', 'lab' & VARCHAR(50) & NOT NULL, CHECK \\
        \hline
        hospital\_id & Reference to the hospital this user belongs to (NULL for patients who may not be associated with a specific hospital) & UUID & FOREIGN KEY references hospitals \\
        \hline
        is\_active & Flag indicating whether the account is active (true) or deactivated/suspended (false) & BOOLEAN & DEFAULT true \\
        \hline
        last\_login & Timestamp of the user's most recent successful login, used for security monitoring and inactivity checks & TIMESTAMP & NULL \\
        \hline
        created\_at & Timestamp when the user account was created, automatically set by database & TIMESTAMP & DEFAULT CURRENT\_TIMESTAMP \\
        \hline
        updated\_at & Timestamp when the user record was last modified, updated automatically via trigger & TIMESTAMP & DEFAULT CURRENT\_TIMESTAMP \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Hospitals Schema - Stores comprehensive information about healthcare facilities registered in the BondHealth platform. Each hospital can have multiple doctors and lab technicians associated with it. The table includes both contact information and classification details for filtering and reporting.}
    \label{tab:hospitals}
    \begin{tabular}{|p{3cm}|p{5cm}|p{3cm}|p{3cm}|}
        \hline
        \textbf{Property Name} & \textbf{Description} & \textbf{Data Type} & \textbf{Constraints} \\
        \hline
        hospital\_id & Unique identifier for each hospital, UUID format for distributed systems & UUID & PRIMARY KEY, DEFAULT uuid\_generate\_v4() \\
        \hline
        hospital\_uuid & Human-readable unique identifier for display purposes, often used in URLs and reports & VARCHAR(50) & UNIQUE \\
        \hline
        name & Official name of the hospital or clinic as registered with healthcare authorities & VARCHAR(255) & NOT NULL \\
        \hline
        type & Classification of healthcare facility: Govt (government hospital), Private (private hospital), Clinic (smaller clinic) & VARCHAR(50) & CHECK \\
        \hline
        address & Complete physical address including street, building, landmark, and postal code & TEXT & NULL \\
        \hline
        city & City where the hospital is located, used for geographic searching and filtering & VARCHAR(100) & NULL \\
        \hline
        phone & Primary contact phone number for the hospital reception or administration & VARCHAR(20) & NULL \\
        \hline
        email & Official email address for hospital communications & VARCHAR(255) & NULL \\
        \hline
        created\_at & Timestamp when the hospital record was added to the system & TIMESTAMP & DEFAULT CURRENT\_TIMESTAMP \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Patients Schema - Extended profile information for users with role='patient'. Contains medical information, emergency contacts, and healthcare-specific data fields. The use of array data types (TEXT[]) allows efficient storage of multiple medical conditions and allergies without creating separate junction tables, simplifying queries and improving performance.}
    \label{tab:patients}
    \begin{tabular}{|p{3cm}|p{5cm}|p{3cm}|p{3cm}|}
        \hline
        \textbf{Property Name} & \textbf{Description} & \textbf{Data Type} & \textbf{Constraints} \\
        \hline
        patient\_id & Unique identifier for patient record, separate from user\_id to allow multiple patient profiles per user (e.g., family accounts for parents managing children) & UUID & PRIMARY KEY, DEFAULT uuid\_generate\_v4() \\
        \hline
        user\_id & Reference to the associated user account in users table, enables one-to-one relationship between users and patients & UUID & UNIQUE, FOREIGN KEY references users ON DELETE CASCADE \\
        \hline
        patient\_uuid & Display identifier for patient, often used on reports and in communications & VARCHAR(50) & UNIQUE \\
        \hline
        full\_name & Patient's complete legal name as per identification documents & VARCHAR(255) & NOT NULL \\
        \hline
        date\_of\_birth & Patient's date of birth for age calculation and medical context & DATE & NULL \\
        \hline
        gender & Patient's gender identity for medical records & VARCHAR(20) & NULL \\
        \hline
        blood\_type & Blood group classification (A+, A-, B+, B-, AB+, AB-, O+, O-) for emergency reference & VARCHAR(5) & NULL \\
        \hline
        phone & Primary contact number for the patient & VARCHAR(20) & NULL \\
        \hline
        email & Patient's email address for communications & VARCHAR(255) & NULL \\
        \hline
        address & Residential address of the patient & TEXT & NULL \\
        \hline
        emergency\_contact\_name & Name of person to contact in case of emergency & VARCHAR(255) & NULL \\
        \hline
        emergency\_contact\_phone & Phone number of emergency contact person & VARCHAR(20) & NULL \\
        \hline
        medical\_conditions & Array of diagnosed medical conditions, stored as text array for efficient querying and filtering & TEXT[] & NULL \\
        \hline
        allergies & Array of known allergies and adverse reactions, stored as text array & TEXT[] & NULL \\
        \hline
        last\_visit & Date of patient's most recent visit to any doctor in the network, used for engagement tracking & DATE & NULL \\
        \hline
        next\_appointment & Date of patient's next scheduled appointment (denormalized for dashboard performance and quick display) & DATE & NULL \\
        \hline
        created\_at & Timestamp when patient record was created & TIMESTAMP & DEFAULT CURRENT\_TIMESTAMP \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Doctors Schema - Comprehensive profile information for healthcare providers. Includes professional details, availability patterns, and performance metrics. The \texttt{available\_days} array stores days of the week when the doctor practices, enabling flexible scheduling patterns. The rating is calculated from patient feedback.}
    \label{tab:doctors}
    \begin{tabular}{|p{3cm}|p{5cm}|p{3cm}|p{3cm}|}
        \hline
        \textbf{Property Name} & \textbf{Description} & \textbf{Data Type} & \textbf{Constraints} \\
        \hline
        doctor\_id & Unique identifier for doctor record & UUID & PRIMARY KEY, DEFAULT uuid\_generate\_v4() \\
        \hline
        user\_id & Reference to associated user account, enabling login capabilities and linking to users table & UUID & UNIQUE, FOREIGN KEY references users ON DELETE CASCADE \\
        \hline
        hospital\_id & Reference to hospital where doctor primarily practices & UUID & FOREIGN KEY references hospitals \\
        \hline
        doctor\_uuid & Display identifier for doctor, used in URLs and reports & VARCHAR(50) & UNIQUE \\
        \hline
        full\_name & Doctor's full name as per medical registration & VARCHAR(255) & NOT NULL \\
        \hline
        specialization & Primary medical specialty (e.g., Cardiology, Pediatrics, Orthopedics) & VARCHAR(100) & NULL \\
        \hline
        designation & Job title or position (e.g., Senior Consultant, Head of Department, Resident) & VARCHAR(100) & NULL \\
        \hline
        qualification & Educational qualifications and certifications (e.g., MBBS, MD, DM, Fellowship) & VARCHAR(255) & NULL \\
        \hline
        experience & Years of professional experience, stored as text to allow ranges (e.g., ``10+ years'', ``5-10 years'') & VARCHAR(50) & NULL \\
        \hline
        consultation\_fee & Standard fee for a consultation, stored as text to accommodate currency symbols and different fee structures & VARCHAR(50) & NULL \\
        \hline
        email & Professional email address for communications & VARCHAR(255) & NULL \\
        \hline
        phone & Contact number for the doctor's office or personal (if provided) & VARCHAR(20) & NULL \\
        \hline
        address & Clinic or hospital address where doctor sees patients & TEXT & NULL \\
        \hline
        photo\_url & URL to doctor's profile photograph stored in cloud storage & TEXT & NULL \\
        \hline
        available\_days & Array of days when doctor is available (e.g., \{'Monday','Wednesday','Friday'\}) & TEXT[] & NULL \\
        \hline
        available\_time & Text description of working hours (e.g., ``9:00 AM - 5:00 PM'', ``10:00 AM - 2:00 PM'') & VARCHAR(100) & NULL \\
        \hline
        rating & Average patient rating calculated from feedback, scale 0-5 with two decimal places & DECIMAL(3,2) & DEFAULT 0 \\
        \hline
        status & Current availability status: 'Available', 'Busy', 'On Leave', 'Offline' for real-time indicators & VARCHAR(50) & DEFAULT 'Available' \\
        \hline
        created\_at & Timestamp when doctor record was created & TIMESTAMP & DEFAULT CURRENT\_TIMESTAMP \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Appointments Schema - Core transactional table recording all patient-doctor appointments. Tracks the complete lifecycle from booking through completion or cancellation. Includes status tracking, timing information, and audit fields. Each appointment is uniquely identified and linked to patient, doctor, and hospital.}
    \label{tab:appointments}
    \begin{tabular}{|p{3cm}|p{5cm}|p{3cm}|p{3cm}|}
        \hline
        \textbf{Property Name} & \textbf{Description} & \textbf{Data Type} & \textbf{Constraints} \\
        \hline
        appointment\_id & Unique identifier for each appointment & UUID & PRIMARY KEY, DEFAULT uuid\_generate\_v4() \\
        \hline
        appointment\_uuid & Human-readable reference number for appointments, used in communications and patient portals & VARCHAR(50) & UNIQUE \\
        \hline
        patient\_id & Reference to the patient who booked the appointment & UUID & FOREIGN KEY references patients ON DELETE CASCADE \\
        \hline
        doctor\_id & Reference to the doctor who will provide the consultation & UUID & FOREIGN KEY references doctors ON DELETE CASCADE \\
        \hline
        hospital\_id & Reference to the hospital where appointment will take place & UUID & FOREIGN KEY references hospitals \\
        \hline
        appointment\_date & Date of the scheduled appointment & DATE & NOT NULL \\
        \hline
        appointment\_time & Time slot for the appointment (e.g., ``10:30 AM'', ``14:00'') & VARCHAR(20) & NOT NULL \\
        \hline
        reason & Patient's reason for visit or chief complaint, captured during booking & TEXT & NULL \\
        \hline
        status & Current status: 'pending', 'confirmed', 'completed', 'cancelled', 'rescheduled', 'no-show' for lifecycle tracking & VARCHAR(50) & DEFAULT 'pending' \\
        \hline
        type & Appointment type: 'in-person', 'video', 'phone' for distinguishing consultation modes & VARCHAR(20) & DEFAULT 'in-person' \\
        \hline
        token\_number & Queue token or sequence number for physical appointments & VARCHAR(20) & NULL \\
        \hline
        location & Specific room or location within hospital (e.g., ``Room 105, 2nd Floor'') & VARCHAR(255) & NULL \\
        \hline
        notes & Internal notes about the appointment (not visible to patient), used by staff & TEXT & NULL \\
        \hline
        created\_at & Timestamp when appointment was booked & TIMESTAMP & DEFAULT CURRENT\_TIMESTAMP \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Lab Reports Schema - Stores all laboratory test results and related information. Supports file attachments, priority levels, and sharing preferences. Enables complete tracking of the lab workflow from test order to result delivery, with appropriate metadata for clinical context.}
    \label{tab:labreports}
    \begin{tabular}{|p{3cm}|p{5cm}|p{3cm}|p{3cm}|}
        \hline
        \textbf{Property Name} & \textbf{Description} & \textbf{Data Type} & \textbf{Constraints} \\
        \hline
        report\_id & Unique identifier for each lab report & UUID & PRIMARY KEY, DEFAULT uuid\_generate\_v4() \\
        \hline
        report\_uuid & Display reference number for reports, used in patient portals and communications & VARCHAR(50) & UNIQUE \\
        \hline
        patient\_id & Reference to the patient who is the subject of the test & UUID & FOREIGN KEY references patients ON DELETE CASCADE \\
        \hline
        doctor\_id & Reference to the doctor who ordered the test & UUID & FOREIGN KEY references doctors \\
        \hline
        lab\_tech\_id & Reference to the lab technician who processed the test and entered results & UUID & FOREIGN KEY references lab\_technicians \\
        \hline
        test\_type & Type of laboratory test performed (e.g., ``Complete Blood Count'', ``Lipid Profile'', ``Blood Glucose'') & VARCHAR(100) & NOT NULL \\
        \hline
        test\_date & Date when the test was performed (sample collection date) & DATE & NULL \\
        \hline
        results & Detailed test results, may include structured data or free text depending on test type & TEXT & NULL \\
        \hline
        findings & Doctor's interpretation and clinical findings after reviewing results & TEXT & NULL \\
        \hline
        file\_url & URL to uploaded report file (PDF, image) in cloud storage & TEXT & NULL \\
        \hline
        status & Report status: 'pending', 'completed', 'reviewed', 'archived' for workflow tracking & VARCHAR(50) & DEFAULT 'pending' \\
        \hline
        priority & Urgency level: 'normal', 'urgent', 'critical' affecting notification behavior and display & VARCHAR(20) & DEFAULT 'normal' \\
        \hline
        shared\_with & Access control: 'doctor', 'patient', 'both' determining who can view the report & VARCHAR(20) & DEFAULT 'doctor' \\
        \hline
        created\_at & Timestamp when report was created (entry time) & TIMESTAMP & DEFAULT CURRENT\_TIMESTAMP \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Prescriptions Schema - Manages medication prescriptions issued by doctors. Supports active prescription tracking, refill management, and medication history. Each prescription record represents a single medication with dosage instructions, allowing for multiple medications per consultation.}
    \label{tab:prescriptions}
    \begin{tabular}{|p{3cm}|p{5cm}|p{3cm}|p{3cm}|}
        \hline
        \textbf{Property Name} & \textbf{Description} & \textbf{Data Type} & \textbf{Constraints} \\
        \hline
        prescription\_id & Unique identifier for each prescription item & UUID & PRIMARY KEY, DEFAULT uuid\_generate\_v4() \\
        \hline
        prescription\_uuid & Reference number for the prescription, groups related medications from the same consultation & VARCHAR(50) & UNIQUE \\
        \hline
        patient\_id & Reference to the patient receiving the medication & UUID & FOREIGN KEY references patients ON DELETE CASCADE \\
        \hline
        doctor\_id & Reference to the prescribing doctor & UUID & FOREIGN KEY references doctors \\
        \hline
        medicine\_name & Name of the prescribed medication (generic or brand name) & VARCHAR(255) & NOT NULL \\
        \hline
        dosage & Dosage amount and form (e.g., ``500mg'', ``1 tablet'', ``10ml'', ``2 puffs'') & VARCHAR(100) & NULL \\
        \hline
        frequency & Administration schedule (e.g., ``Twice daily'', ``Every 8 hours'', ``Once daily at bedtime'') & VARCHAR(255) & NULL \\
        \hline
        valid\_until & Date until which the prescription remains valid for refills & DATE & NULL \\
        \hline
        instructions & Additional instructions for the patient (e.g., ``Take with food'', ``Avoid alcohol'') & TEXT & NULL \\
        \hline
        status & Prescription status: 'active', 'completed', 'expired', 'cancelled' for lifecycle tracking & VARCHAR(50) & DEFAULT 'active' \\
        \hline
        created\_at & Timestamp when prescription was written & TIMESTAMP & DEFAULT CURRENT\_TIMESTAMP \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Doctor Leave Schema - Tracks leave applications and approvals for doctors. Ensures appointment availability reflects leave periods and enables coverage planning. Supports date ranges and status tracking for approval workflow.}
    \label{tab:doctorleave}
    \begin{tabular}{|p{3cm}|p{5cm}|p{3cm}|p{3cm}|}
        \hline
        \textbf{Property Name} & \textbf{Description} & \textbf{Data Type} & \textbf{Constraints} \\
        \hline
        leave\_id & Unique identifier for each leave application & UUID & PRIMARY KEY, DEFAULT uuid\_generate\_v4() \\
        \hline
        doctor\_id & Reference to the doctor applying for leave & UUID & FOREIGN KEY references doctors ON DELETE CASCADE \\
        \hline
        from\_date & Start date of leave period (inclusive) & DATE & NOT NULL \\
        \hline
        to\_date & End date of leave period (inclusive) & DATE & NOT NULL \\
        \hline
        reason & Reason for leave (e.g., ``Vacation'', ``Sick leave'', ``Conference'', ``Personal'') & TEXT & NULL \\
        \hline
        status & Leave status: 'Pending', 'Approved', 'Rejected', 'Completed', 'Cancelled' for workflow & VARCHAR(50) & DEFAULT 'Approved' \\
        \hline
        created\_at & Timestamp when leave was applied & TIMESTAMP & DEFAULT CURRENT\_TIMESTAMP \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Medicines Inventory Schema - Manages hospital pharmacy inventory, tracking stock levels, pricing, and expiry dates. Enables automated reorder alerts and expiry notifications to ensure medication availability and patient safety.}
    \label{tab:medicines}
    \begin{tabular}{|p{3cm}|p{5cm}|p{3cm}|p{3cm}|}
        \hline
        \textbf{Property Name} & \textbf{Description} & \textbf{Data Type} & \textbf{Constraints} \\
        \hline
        medicine\_id & Unique identifier for each medicine item in inventory & UUID & PRIMARY KEY, DEFAULT uuid\_generate\_v4() \\
        \hline
        hospital\_id & Reference to the hospital that maintains this inventory & UUID & FOREIGN KEY references hospitals \\
        \hline
        name & Name of the medicine (brand name or generic) with strength & VARCHAR(255) & NOT NULL \\
        \hline
        quantity & Current stock quantity available (number of units) & INTEGER & DEFAULT 0 \\
        \hline
        price & Price per unit, stored as text to accommodate currency formatting and different pricing models & VARCHAR(50) & NULL \\
        \hline
        expiry\_date & Date when the medicine batch expires, used for alerts and stock rotation & DATE & NULL \\
        \hline
        created\_at & Timestamp when medicine record was added to inventory & TIMESTAMP & DEFAULT CURRENT\_TIMESTAMP \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Payments Schema - Records all financial transactions processed through the platform. Links to appointments for context, tracks payment status, and stores gateway transaction references for reconciliation. Supports multiple payment methods and currencies.}
    \label{tab:payments}
    \begin{tabular}{|p{3cm}|p{5cm}|p{3cm}|p{3cm}|}
        \hline
        \textbf{Property Name} & \textbf{Description} & \textbf{Data Type} & \textbf{Constraints} \\
        \hline
        payment\_id & Unique identifier for each payment transaction & UUID & PRIMARY KEY, DEFAULT uuid\_generate\_v4() \\
        \hline
        payment\_uuid & Reference number for payments, used on receipts and in communications & VARCHAR(50) & UNIQUE \\
        \hline
        appointment\_id & Reference to the appointment being paid for (if applicable) & UUID & FOREIGN KEY references appointments \\
        \hline
        patient\_id & Reference to the patient making the payment & UUID & FOREIGN KEY references patients \\
        \hline
        amount & Monetary amount of the transaction with two decimal precision & DECIMAL(10,2) & NOT NULL \\
        \hline
        currency & Three-letter currency code (e.g., 'USD', 'EUR', 'INR', 'GBP') & VARCHAR(10) & DEFAULT 'USD' \\
        \hline
        status & Payment status: 'pending', 'completed', 'failed', 'refunded', 'partially\_refunded' for tracking & VARCHAR(50) & DEFAULT 'pending' \\
        \hline
        payment\_method & Method used: 'card', 'cash', 'insurance', 'wallet', 'bank\_transfer' for reconciliation & VARCHAR(50) & NULL \\
        \hline
        transaction\_id & External transaction ID from payment gateway for reconciliation and support & VARCHAR(255) & NULL \\
        \hline
        invoice\_url & URL to generated invoice PDF stored in cloud storage & TEXT & NULL \\
        \hline
        created\_at & Timestamp when payment was initiated & TIMESTAMP & DEFAULT CURRENT\_TIMESTAMP \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Messages Schema (Real-time Chat) - Stores all chat communications between patients and doctors. Supports text messages and file attachments, with read receipts for delivery tracking. Used for both real-time and asynchronous communication, maintaining complete conversation history for medical-legal purposes.}
    \label{tab:messages}
    \begin{tabular}{|p{3cm}|p{5cm}|p{3cm}|p{3cm}|}
        \hline
        \textbf{Property Name} & \textbf{Description} & \textbf{Data Type} & \textbf{Constraints} \\
        \hline
        message\_id & Unique identifier for each message & UUID & PRIMARY KEY, DEFAULT uuid\_generate\_v4() \\
        \hline
        sender\_id & Reference to the user who sent the message & UUID & FOREIGN KEY references users \\
        \hline
        receiver\_id & Reference to the user who should receive the message & UUID & FOREIGN KEY references users \\
        \hline
        content & Text content of the message (can be NULL if only attachment) & TEXT & NULL \\
        \hline
        attachment\_url & URL to attached file (images, documents) if any & TEXT & NULL \\
        \hline
        is\_read & Flag indicating whether the message has been read by the receiver & BOOLEAN & DEFAULT false \\
        \hline
        created\_at & Timestamp when message was sent & TIMESTAMP & DEFAULT CURRENT\_TIMESTAMP \\
        \hline
    \end{tabular}
\end{table}

\textbf{Database Indexes for Performance Optimization:}

To ensure optimal query performance as the database grows to millions of records, the following indexes have been created on frequently queried columns. These indexes significantly reduce query execution time by allowing the database to quickly locate relevant rows without scanning entire tables.

\begin{itemize}
    \item \texttt{CREATE INDEX idx\_users\_email ON users(email);} - Accelerates login and user lookup by email, one of the most frequent operations
    \item \texttt{CREATE INDEX idx\_users\_username ON users(username);} - Speeds up login by username for users who prefer username authentication
    \item \texttt{CREATE INDEX idx\_patients\_user\_id ON patients(user\_id);} - Fast patient profile access from user ID during authentication
    \item \texttt{CREATE INDEX idx\_doctors\_hospital\_id ON doctors(hospital\_id);} - Quick doctor lookup by hospital for directory features
    \item \texttt{CREATE INDEX idx\_doctors\_specialization ON doctors(specialization);} - Enables fast doctor filtering by specialty for patient searches
    \item \texttt{CREATE INDEX idx\_appointments\_patient\_id ON appointments(patient\_id);} - Rapid retrieval of patient appointment history for dashboards
    \item \texttt{CREATE INDEX idx\_appointments\_doctor\_id ON appointments(doctor\_id);} - Fast doctor schedule access for availability checking
    \item \texttt{CREATE INDEX idx\_appointments\_date ON appointments(appointment\_date);} - Optimizes date-based queries for daily schedules and reports
    \item \texttt{CREATE INDEX idx\_lab\_reports\_patient\_id ON lab\_reports(patient\_id);} - Quick patient report history access for medical records
    \item \texttt{CREATE INDEX idx\_lab\_reports\_status ON lab\_reports(status);} - Enables fast filtering by report status for lab dashboards
    \item \texttt{CREATE INDEX idx\_prescriptions\_patient\_id ON prescriptions(patient\_id);} - Rapid prescription history retrieval for patient records
    \item \texttt{CREATE INDEX idx\_messages\_sender\_receiver ON messages(sender\_id, receiver\_id);} - Optimizes chat conversation queries for message history
    \item \texttt{CREATE INDEX idx\_messages\_created\_at ON messages(created\_at DESC);} - Speeds up message history display in chronological order
\end{itemize}

% =============================================================================
% 4.5 API DESIGN
% =============================================================================

\section{API Design (RESTful Endpoints)}

The BondHealth platform implements a comprehensive RESTful API architecture that follows industry best practices for web service design. All endpoints adhere to REST conventions with appropriate HTTP methods, consistent resource naming, and standardized response formats. The API is designed to be intuitive, self-documenting, and maintainable while ensuring security through authentication middleware and role-based access controls.

\textbf{API Design Principles:}

\begin{itemize}
    \item \textbf{Resource-Oriented}: Endpoints are organized around resources (patients, doctors, appointments) rather than actions, making the API intuitive and consistent with REST philosophy.
    
    \item \textbf{Consistent Naming}: Plural nouns for collections (/appointments) and singular for specific resources (/appointments/\{id\}) provide clear and predictable URL patterns.
    
    \item \textbf{HTTP Methods}: GET for retrieval, POST for creation, PUT for full updates, PATCH for partial updates, DELETE for removal - each method has a specific, well-defined purpose.
    
    \item \textbf{Status Codes}: Appropriate HTTP status codes (200 OK, 201 Created, 400 Bad Request, 401 Unauthorized, 403 Forbidden, 404 Not Found, 500 Internal Server Error) provide clear feedback about operation outcomes.
    
    \item \textbf{Versioning}: API versioning through URL path (/api/v1/...) supports future changes without breaking existing clients.
    
    \item \textbf{Response Format}: Consistent JSON structure with success/data/error fields makes client-side handling predictable and straightforward.
    
    \item \textbf{Authentication}: JWT tokens in HTTP-only cookies for all protected endpoints, ensuring secure, automatic token handling.
    
    \item \textbf{Rate Limiting}: Configurable rate limits per endpoint prevent abuse and ensure fair resource allocation across all users.
\end{itemize}

The following tables document all API endpoints organized by functional area. Each endpoint includes the HTTP method, URL path, description, and access level required.

\begin{table}[h]
    \centering
    \caption{Authentication Endpoints - Public endpoints for user registration, login, and session management. These endpoints handle identity verification, token issuance, and session lifecycle. All authentication endpoints use HTTPS exclusively to prevent credential interception.}
    \label{tab:auth-endpoints}
    \begin{tabular}{|p{2cm}|p{3.5cm}|p{4cm}|p{3cm}|}
        \hline
        \textbf{Method} & \textbf{Endpoint} & \textbf{Description} & \textbf{Access} \\
        \hline
        POST & /api/auth/register & Register new user account with email verification. Accepts user details, validates input, creates user record, and sends verification email. Returns success with user ID and prompts for email verification. & Public \\
        \hline
        POST & /api/auth/login & Authenticate user credentials. Validates username/email and password, generates JWT token, sets HTTP-only cookie, records last login, and returns user profile with role-specific redirect URL. Implements rate limiting to prevent brute force. & Public \\
        \hline
        POST & /api/auth/logout & Terminate user session. Clears authentication cookie, removes session from active sessions store, and logs logout event for audit. & Authenticated \\
        \hline
        GET & /api/auth/me & Retrieve currently authenticated user's profile information. Validates token, fetches fresh user data from database, and returns complete profile including role-specific details. Used for session validation and profile display. & Authenticated \\
        \hline
        POST & /api/auth/refresh & Refresh expired JWT token. Validates refresh token, issues new access token, and updates cookie. Enables long-lived sessions without requiring re-login. & Authenticated \\
        \hline
        POST & /api/auth/forgot-password & Initiate password reset process. Sends password reset email with secure, time-limited token to user's registered email. & Public \\
        \hline
        POST & /api/auth/reset-password & Complete password reset using token from email. Validates token, updates password with new hash, and invalidates all existing sessions. & Public (with token) \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Patient Endpoints - Comprehensive set of endpoints for patient-facing functionality. All endpoints require patient role authentication and provide access to patient-specific data. Responses are filtered to include only information relevant to the authenticated patient.}
    \label{tab:patient-endpoints}
    \begin{tabular}{|p{2cm}|p{3.5cm}|p{4cm}|p{3cm}|}
        \hline
        \textbf{Method} & \textbf{Endpoint} & \textbf{Description} & \textbf{Access} \\
        \hline
        GET & /api/patient & Retrieve complete patient profile including personal information, medical conditions, allergies, emergency contacts, and notification preferences. Joins data from users and patients tables for comprehensive response. & Patient \\
        \hline
        PUT & /api/patient & Update patient profile information. Validates input, updates allowed fields, and returns updated profile. Supports partial updates via JSON merge patch. & Patient \\
        \hline
        GET & /api/patient/medical-history & Retrieve complete medical history including past appointments, diagnoses, procedures, and clinical notes in chronological order. Supports date range filtering. & Patient \\
        \hline
        GET & /api/appointments & Get paginated list of patient's appointments with filtering by status (upcoming/past/completed), date range, and doctor. Returns appointment details including doctor information and location. & Patient \\
        \hline
        GET & /api/appointments/:id & Get detailed information for a specific appointment including full doctor profile, location details, payment status, and any consultation notes (if completed). & Patient \\
        \hline
        POST & /api/appointments & Book new appointment. Accepts doctor\_id, date, time, reason, type. Validates availability, creates appointment record with pending status, triggers payment if required, and returns created appointment with confirmation details. & Patient \\
        \hline
        PUT & /api/appointments/:id/reschedule & Reschedule existing appointment. Accepts new\_date, new\_time, reason. Validates new slot availability, updates appointment with rescheduled status, maintains original for audit, and notifies doctor. & Patient \\
        \hline
        PUT & /api/appointments/:id/cancel & Cancel appointment. Accepts cancellation reason. Updates status to 'cancelled', processes refund if applicable based on policy, and sends notifications to all affected parties. & Patient \\
        \hline
        GET & /api/reports & Get paginated list of patient's lab reports. Supports filtering by date range, test type, and status. Returns report metadata and secure access URLs with expiration. & Patient \\
        \hline
        GET & /api/reports/:id/download & Download specific lab report file. Validates access permissions, generates temporary signed URL with 5-minute validity, and redirects to file in cloud storage. & Patient \\
        \hline
        GET & /api/prescriptions & Get list of patient's prescriptions with filtering by active/expired status. Returns prescription details, medicine information, and refill eligibility. & Patient \\
        \hline
        POST & /api/prescriptions/:id/refill & Request refill for an active prescription. Validates refill eligibility (refills remaining, not expired), creates refill request notification for doctor, and tracks request status. & Patient \\
        \hline
        GET & /api/doctors & Search and filter available doctors. Supports filtering by specialization, hospital, availability, ratings, languages, and insurance accepted. Returns paginated results with availability summary. & Patient \\
        \hline
        GET & /api/doctors/:id/availability & Get available time slots for a specific doctor on a given date. Returns list of available slots based on schedule, existing appointments, and leave periods. Implements slot locking during selection. & Patient \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Doctor Endpoints - Endpoints for doctor-specific functionality. All endpoints require doctor role authentication and provide access to patient data and clinical tools. Responses include only patients assigned to or seen by the authenticated doctor.}
    \label{tab:doctor-endpoints}
    \begin{tabular}{|p{2cm}|p{3.5cm}|p{4cm}|p{3cm}|}
        \hline
        \textbf{Method} & \textbf{Endpoint} & \textbf{Description} & \textbf{Access} \\
        \hline
        GET & /api/doctor & Retrieve doctor's complete profile including professional information, schedule, consultation fees, performance metrics, and practice settings. & Doctor \\
        \hline
        PUT & /api/doctor & Update doctor profile. Allows modification of consultation fees, availability schedule, qualifications, contact information, and practice settings. Changes may require admin approval. & Doctor \\
        \hline
        GET & /api/appointments/today & Get today's complete appointment schedule ordered by time. Includes patient information, appointment status, and quick action flags. Updates in real-time as appointments change. & Doctor \\
        \hline
        GET & /api/appointments/upcoming & Get future appointments beyond today. Supports pagination, filtering by date range, and sorting. Returns comprehensive appointment details with patient summaries. & Doctor \\
        \hline
        GET & /api/appointments/:id & Get detailed appointment information including full patient history, previous consultations, and any pre-visit questionnaires completed. & Doctor \\
        \hline
        PUT & /api/appointments/:id/start & Mark appointment as started when consultation begins. Updates status to 'in-progress', records start time, and triggers any time-tracking logic. & Doctor \\
        \hline
        PUT & /api/appointments/:id/complete & Mark appointment as completed after consultation. Updates status to 'completed', records end time, calculates duration, and triggers follow-up actions like feedback requests. & Doctor \\
        \hline
        GET & /api/patients & Get paginated list of doctor's patients. Supports search by name, filtering by recent visits, conditions, and sorting. Returns patient summaries with key information. & Doctor \\
        \hline
        GET & /api/patients/:id & Get complete patient profile including medical history, conditions, allergies, past consultations with this doctor, and recent lab reports. & Doctor \\
        \hline
        GET & /api/lab-reports & Get pending lab reports requiring review. Supports filtering by priority (normal/urgent/critical) and date range. Returns reports needing clinical findings with patient context. & Doctor \\
        \hline
        GET & /api/lab-reports/:id & Get complete lab report details including test results, reference ranges, patient information, and any existing findings. Includes historical comparison when available. & Doctor \\
        \hline
        PUT & /api/lab-reports/:id/findings & Add clinical findings and interpretation to lab report. Accepts findings text, updates report status to 'reviewed', and notifies relevant parties (patient if shared). & Doctor \\
        \hline
        POST & /api/prescriptions & Create new prescription for patient. Accepts medicine details, dosage, frequency, duration, and instructions. Validates against patient allergies and potential drug interactions. & Doctor \\
        \hline
        PUT & /api/prescriptions/:id & Update existing prescription (dosage adjustments, renewals, discontinuation). Maintains version history and requires reason for changes. & Doctor \\
        \hline
        POST & /api/doctors/:doctorId/leave & Apply for leave. Accepts from\_date, to\_date, reason. Creates leave request with pending status for admin approval. Automatically blocks appointments for the period upon approval. & Doctor \\
        \hline
        GET & /api/doctors/leave/status & View doctor's leave requests and approval status. Shows upcoming approved leaves and pending requests with expected response times. & Doctor \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Admin Endpoints - Administrative endpoints for hospital management. All endpoints require admin role authentication and provide comprehensive control over hospital operations, user management, and system configuration.}
    \label{tab:admin-endpoints}
    \begin{tabular}{|p{2cm}|p{3.5cm}|p{4cm}|p{3cm}|}
        \hline
        \textbf{Method} & \textbf{Endpoint} & \textbf{Description} & \textbf{Access} \\
        \hline
        GET & /api/hospital/data & Get comprehensive hospital data including statistics, doctor counts, appointment volumes, inventory status, and financial summaries. Returns dashboard data with visual-ready metrics. & Admin \\
        \hline
        PUT & /api/hospital/profile & Update hospital profile information including address, contact details, operating hours, branding, and configuration settings. Changes are audited. & Admin \\
        \hline
        POST & /api/hospital/add/doctor & Register new doctor in the system. Accepts complete doctor profile, creates user account with doctor role, assigns to departments, and sends welcome email with credentials. & Admin \\
        \hline
        PUT & /api/hospital/doctors/:id & Update doctor information including profile, fees, schedule, department assignment, and employment status. Maintains change history. & Admin \\
        \hline
        DELETE & /api/hospital/doctors/:id & Deactivate or remove doctor from system. Handles reassignment of future appointments to other doctors if needed and archives records appropriately. & Admin \\
        \hline
        GET & /api/hospital/doctors & Get paginated list of all doctors with filtering by department, specialization, status, and search. Returns comprehensive profiles with performance metrics. & Admin \\
        \hline
        POST & /api/hospital/add/medicine & Add new medicine to inventory. Accepts medicine name, initial quantity, price, expiry date, manufacturer, and storage requirements. & Admin \\
        \hline
        PUT & /api/hospital/medicines/:id & Update medicine information including quantity adjustments (with reason), price changes, expiry updates, and reorder levels. & Admin \\
        \hline
        GET & /api/hospital/medicines/expiring & Get list of medicines expiring within configurable date range for alerts and disposal planning. Includes quantity and location. & Admin \\
        \hline
        GET & /api/hospital/medicines/low-stock & Get medicines below reorder level for replenishment planning. Includes current quantity, reorder level, and suggested order quantity. & Admin \\
        \hline
        POST & /api/hospital/add/lab & Add new lab technician. Creates user account with lab role, assigns to hospital and specific lab/department, and configures permissions. & Admin \\
        \hline
        GET & /api/doctors/on-leave & Get list of doctors currently on leave or with approved future leave. Includes leave details and coverage arrangements if any. & Admin \\
        \hline
        GET & /api/leave/requests & Get all pending leave requests requiring approval. Supports filtering by doctor, department, date range. Returns requests with doctor details. & Admin \\
        \hline
        POST & /api/doctors/:doctorId/leave/:leaveId/approve & Approve pending leave request. Updates status to 'approved', blocks appointments for the period in doctor's schedule, and notifies doctor. & Admin \\
        \hline
        POST & /api/doctors/:doctorId/leave/:leaveId/reject & Reject leave request with reason. Updates status to 'rejected', sends notification to doctor with reason, and maintains record for audit. & Admin \\
        \hline
        GET & /api/appointments/all & Get all hospital appointments with comprehensive filtering (date range, status, doctor, department) and pagination. Used for operational oversight. & Admin \\
        \hline
        GET & /api/reports/financial & Generate financial reports for specified period including revenue by department, payment collections, refunds, and outstanding balances. Supports multiple formats. & Admin \\
        \hline
        GET & /api/reports/analytics & Get operational analytics including appointment volumes, no-show rates, popular services, patient demographics, and doctor performance. & Admin \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Lab Technician Endpoints - Endpoints for laboratory workflow management. All endpoints require lab role authentication and provide tools for test processing, result entry, and report management.}
    \label{tab:labtech-endpoints}
    \begin{tabular}{|p{2cm}|p{3.5cm}|p{4cm}|p{3cm}|}
        \hline
        \textbf{Method} & \textbf{Endpoint} & \textbf{Description} & \textbf{Access} \\
        \hline
        GET & /api/lab/dashboard & Get lab technician dashboard data including today's test counts, pending items by priority, completed tests, urgent cases requiring immediate attention, and recent activity. & Lab \\
        \hline
        POST & /api/lab-reports & Create new lab report. Accepts patient\_id, doctor\_id, test\_type, results, priority, sharing preferences, and file uploads. Validates required fields, performs range checks, and stores report with appropriate metadata. & Lab \\
        \hline
        GET & /api/lab-reports/pending & Get list of pending reports requiring completion or review. Supports filtering by priority (normal/urgent/critical) and date. Returns reports with patient and doctor information. & Lab \\
        \hline
        GET & /api/lab-reports/:id & Get specific lab report details including all fields, file references, and processing history. Used for review and editing. & Lab \\
        \hline
        PUT & /api/lab-reports/:id & Update existing lab report (for corrections or additions). Maintains version history and audit trail of changes with before/after values. & Lab \\
        \hline
        DELETE & /api/lab-reports/:id & Delete draft or incorrect report (with appropriate permissions). Only allowed within time window and requires reason. Maintains audit log of deletion. & Lab \\
        \hline
        GET & /api/patient-history & Search patient history for context. Returns recent tests and results for the patient to assist with result correlation and quality control. & Lab \\
        \hline
        GET & /api/patients/search & Search patients by name, ID, or contact for report creation. Returns matching patient summaries with identification details for verification. & Lab \\
        \hline
        POST & /api/lab-reports/:id/upload & Upload additional files to existing report. Accepts multiple files with validation (type, size, virus scan). Updates report with file references. & Lab \\
        \hline
        GET & /api/lab-reports/statistics & Get laboratory statistics including test volumes by type, turnaround times, quality metrics, and technician performance. Used for lab management. & Lab \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Payment Endpoints - Endpoints for payment processing and financial transactions. Access varies by role, with patients able to view their own payments and admins having full access for reconciliation and management.}
    \label{tab:payment-endpoints}
    \begin{tabular}{|p{2cm}|p{3.5cm}|p{4cm}|p{3cm}|}
        \hline
        \textbf{Method} & \textbf{Endpoint} & \textbf{Description} & \textbf{Access} \\
        \hline
        POST & /api/payments/create-intent & Create payment intent for appointment. Returns client secret for front-end tokenization, enabling secure card data capture without exposing sensitive details to server. Used for card payments. & Patient \\
        \hline
        POST & /api/payments/confirm & Confirm payment after gateway processing. Updates payment status from 'pending' to 'completed' or 'failed' based on gateway response, and updates associated appointment status accordingly. & Patient \\
        \hline
        GET & /api/payments/:id & Get detailed payment information including status, amount, method, transaction reference, and timestamps. Used for receipt viewing and reconciliation. & Patient/Admin \\
        \hline
        GET & /api/payments/history & Get paginated payment history for patient (own payments) or all payments (admin). Supports filtering by date range, status, payment method. Returns comprehensive transaction details. & Patient/Admin \\
        \hline
        POST & /api/payments/refund & Process refund for cancelled appointment or adjustment. Requires admin approval for amounts above configurable threshold. Updates payment status and generates credit note. & Admin \\
        \hline
        POST & /api/payments/webhook & Webhook endpoint for payment gateway callbacks. Securely processes asynchronous notifications (payment success/failure, disputes, refunds) and updates internal records accordingly. Endpoint validates gateway signature for security. & Public (secured by signature) \\
        \hline
        GET & /api/payments/invoice/:id & Generate and download invoice PDF for completed payment. Returns PDF file with all required billing details, tax information, and payment confirmation. Cached for performance. & Patient/Admin \\
        \hline
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Chat Endpoints - Endpoints for real-time messaging between patients and doctors. Supports both REST for history retrieval and WebSocket for real-time communication. All messages are encrypted and stored for medical-legal purposes.}
    \label{tab:chat-endpoints}
    \begin{tabular}{|p{2cm}|p{3.5cm}|p{4cm}|p{3cm}|}
        \hline
        \textbf{Method} & \textbf{Endpoint} & \textbf{Description} & \textbf{Access} \\
        \hline
        GET & /api/messages/conversations & Get list of recent conversations for the authenticated user. Includes last message preview, timestamp, unread count, and other participant details. Used for chat list display. & Patient/Doctor \\
        \hline
        GET & /api/messages/:userId & Get complete message history with specific user. Returns paginated messages in chronological order with read status, attachment information, and delivery timestamps. & Patient/Doctor \\
        \hline
        POST & /api/messages & Send new message. Accepts receiver\_id, content, optional attachment. Creates message record in database, triggers real-time delivery via WebSocket to online recipients, and sends push notification to offline users. & Patient/Doctor \\
        \hline
        PUT & /api/messages/:id/read & Mark message as read. Updates is\_read flag and timestamp, triggers read receipt via WebSocket to sender, and updates unread counts. & Patient/Doctor \\
        \hline
        DELETE & /api/messages/:id & Delete message (soft delete with audit trail). Only available within configurable time window (e.g., 1 hour) and requires reason. Message content is replaced with placeholder. & Patient/Doctor \\
        \hline
        GET & /api/messages/unread/count & Get count of unread messages across all conversations for badge notifications. Used for real-time UI updates and push notification triggers. & Patient/Doctor \\
        \hline
    \end{tabular}
\end{table}

% =============================================================================
% 4.6 MODULE-WISE DETAILED DESIGN
% =============================================================================

\section{Module-wise Detailed Design}

This section provides comprehensive design specifications for each major module of the BondHealth platform, describing their internal architecture, component structure, key features, and interactions with other modules. Each module is designed with high cohesion and low coupling, following software engineering best practices to ensure maintainability, testability, and scalability. The modular architecture allows independent development, deployment, and scaling of different functional areas based on usage patterns.

\subsection{Authentication Module}

The authentication module is the security gateway for the BondHealth platform, responsible for verifying user identities, managing sessions, and enforcing access controls. It implements industry-standard security practices while maintaining usability through seamless token management and session handling.

\textbf{Core Components:}

\begin{enumerate}
    \item \textbf{JWT Token Service}: Generates and validates JSON Web Tokens with the following characteristics:
    \begin{itemize}
        \item Tokens contain user ID, username, role, and session ID in the payload for stateless authorization
        \item Signed with HS256 algorithm using a secure server-side secret stored in environment variables
        \item Expiration set to 7 days for access tokens (configurable based on security requirements)
        \item Refresh token mechanism with 30-day validity for extending sessions without re-login
        \item Token versioning (increment on password change) to support forced logout of all sessions
    \end{itemize}
    
    \item \textbf{HTTP-Only Cookie Management}: Tokens are stored in HTTP-only cookies with comprehensive security flags:
    \begin{itemize}
        \item \texttt{HttpOnly}: Prevents JavaScript access, eliminating XSS token theft vectors
        \item \texttt{Secure}: Ensures cookies only sent over HTTPS connections, preventing network interception
        \item \texttt{SameSite=Strict}: Protects against CSRF attacks by restricting cross-origin requests
        \item \texttt{Max-Age}: Sets cookie expiration matching token lifetime for automatic cleanup
        \item Path restriction to API endpoints only, minimizing exposure
    \end{itemize}
    
    \item \textbf{Server-Side Session Store}: Maintains active sessions in memory with Redis backing for distributed deployments:
    \begin{itemize}
        \item Key: session ID (from JWT) for O(1) lookup performance
        \item Value: user ID, creation time, last activity, IP address, user agent for security monitoring
        \item Enables forced logout of specific sessions from admin panel
        \item Session invalidation on password change, account lock, or admin action
        \item Concurrent session limiting (configurable, default 5 per user) to prevent abuse
    \end{itemize}
    
    \item \textbf{Password Hashing Service}: Implements bcrypt with configurable work factor:
    \begin{itemize}
        \item Salt rounds: 10 (balance of security and performance, approximately 100ms per hash)
        \item Automatic salt generation and storage with hash for password verification
        \item Constant-time comparison to prevent timing attacks on password verification
        \item Password complexity validation (minimum 8 chars, mixed case, numbers, special) before hashing
    \end{itemize}
    
    \item \textbf{Authentication Middleware Pipeline}:
    \begin{itemize}
        \item Extract token from cookies (or Authorization header for API clients)
        \item Verify token signature and expiration using JWT library
        \item Validate session existence in session store (prevents use of revoked tokens)
        \item Attach user object to request for downstream use in controllers
        \item Refresh token if approaching expiration (sliding sessions) for active users
        \item Log authentication attempts (success/failure) for security monitoring and audit
    \end{itemize}
    
    \item \textbf{Role-Based Access Control (RBAC) Middleware}:
    \begin{itemize}
        \item Check user.role against required roles for endpoint (array of allowed roles)
        \item Hierarchical role support (admin can access doctor endpoints, but not vice versa)
        \item Permission-based checks for fine-grained control beyond basic roles
        \item Hospital-scoped access for multi-tenant isolation (doctors only see their hospital's data)
    \end{itemize}
\end{enumerate}

\textbf{Authentication Flow:}

\begin{enumerate}
    \item User submits credentials via login form (username/email + password) over HTTPS
    \item Server validates credentials against database:
    \begin{itemize}
        \item Look up user by username/email with case-insensitive matching
        \item Compare provided password with stored hash using bcrypt's constant-time comparison
        \item Check account status (active, not locked, email verified)
        \item Record failed attempts in database and implement rate limiting (5 attempts, 15-minute lockout)
    \end{itemize}
    \item On successful validation:
    \begin{itemize}
        \item Generate JWT with user payload (user\_id, username, role) and unique session ID
        \item Create session record in active sessions store with metadata (IP, user agent)
        \item Set HTTP-only cookie with token and security flags
        \item Update last\_login timestamp in database for audit
        \item Return user profile (sanitized, no password hash) and role-specific redirect URL
    \end{itemize}
    \item On failed validation:
    \begin{itemize}
        \item Increment failed attempt counter for the user
        \item Return 401 Unauthorized with generic error message (not revealing whether username or password was incorrect)
        \item After threshold exceeded, lock account and notify user via email
    \end{itemize}
    \item Subsequent requests include token in cookie automatically by browser
    \item Authentication middleware validates token and session on every protected request
    \item Role-based middleware checks permissions before processing request
    \item Access granted with user context or denied with appropriate HTTP status
\end{enumerate}

\subsection{Patient Dashboard Module}

The patient dashboard provides a comprehensive, intuitive interface for patients to manage all aspects of their healthcare journey. It aggregates information from multiple modules and presents it in a unified view optimized for quick access and clear understanding, following healthcare UX best practices for readability and accessibility.

\textbf{Component Architecture:}

\begin{enumerate}
    \item \textbf{Profile Summary Card}:
    \begin{itemize}
        \item Displays patient name, photo (if available), blood type, and age calculated from DOB
        \item Shows emergency contact information prominently with one-tap dialing on mobile
        \item Includes quick edit button for profile updates (opens modal with editable fields)
        \item Displays insurance information summary (provider, policy number, coverage dates)
        \item Shows account status and verification badges (email verified, phone verified)
    \end{itemize}
    
    \item \textbf{Upcoming Appointments Section}:
    \begin{itemize}
        \item Lists next 3-5 upcoming appointments with date, time, doctor name, and specialty
        \item Each entry includes action buttons: Reschedule, Cancel, View Details
        \item Color-coded status indicators (confirmed in green, pending payment in yellow)
        \item Countdown timer showing time until appointment (days/hours)
        \item Video call join button for telehealth appointments (active 15 minutes before)
        \item Option to add to calendar (Google, Apple, Outlook) with one-click .ics file
    \end{itemize}
    
    \item \textbf{Past Appointments Section}:
    \begin{itemize}
        \item Paginated list of completed appointments with expandable details
        \item Filter by date range, doctor, specialty for easy lookup
        \item Each entry links to consultation summary and doctor's notes
        \item Option to book follow-up appointment with same doctor (pre-filled details)
        \item Feedback and rating interface for completed visits (star rating + comments)
    \end{itemize}
    
    \item \textbf{Medical Records Section}:
    \begin{itemize}
        \item Tabbed interface for different record types with clear organization:
        \begin{itemize}
            \item \textbf{Lab Reports}: List with test type, date, status (new/reviewed), download button
            \item \textbf{Prescriptions}: Active and historical prescriptions with refill status and remaining refills
            \item \textbf{Imaging}: X-rays, MRIs, CT scans with viewer links and radiology reports
            \item \textbf{Vaccinations}: Record of immunizations with due dates for boosters
            \item \textbf{Documents}: Uploaded medical documents (advanced directives, consent forms)
        \end{itemize}
        \item Search and filter capabilities across all records with full-text search
        \item Timeline view showing chronological medical history with visual timeline
        \item Download all records option (ZIP export) for data portability
    \end{itemize}
    
    \item \textbf{Prescription Management Area}:
    \begin{itemize}
        \item Active prescriptions with medicine name, dosage, frequency, and duration
        \item Refill status (refills remaining) and days until expiration
        \item ``Request Refill'' button for eligible prescriptions with one-click action
        \item Medication reminder settings integration (set daily reminders)
        \item Drug interaction warnings displayed prominently if applicable
    \end{itemize}
    
    \item \textbf{Quick Action Buttons}:
    \begin{itemize}
        \item Book New Appointment (opens doctor search modal)
        \item Message My Doctor (opens chat with recent doctors list)
        \item View Lab Reports (jumps to reports section with filters)
        \item Pay Outstanding Bills (opens payment page with due amounts)
        \item Find Nearby Hospitals (location-based search with map)
        \item Emergency Contact (one-tap dialing for emergency numbers)
    \end{itemize}
    
    \item \textbf{Notification Center}:
    \begin{itemize}
        \item Recent notifications with read/unread status and timestamps
        \item Appointment reminders with confirm/decline actions
        \item Lab report ready alerts with direct link to view
        \item Prescription refill reminders with request button
        \item System announcements from hospital administration
        \item Mark all as read functionality and notification preferences link
    \end{itemize}
    
    \item \textbf{Health Metrics Widget} (optional integration):
    \begin{itemize}
        \item Recent vital signs if connected to wearables (heart rate, blood pressure, steps)
        \item Medication adherence tracking with calendar view
        \item Appointment adherence statistics (show/no-show history)
        \item Health goals progress with visual indicators
    \end{itemize}
\end{enumerate}

\textbf{Data Flow:}
\begin{itemize}
    \item Dashboard loads by making parallel API calls to multiple endpoints for performance:
    \begin{itemize}
        \item GET /api/patient for profile (includes medical conditions and emergency contacts)
        \item GET /api/appointments?status=upcoming\&limit=5 for upcoming appointments
        \item GET /api/reports?limit=5\&status=new for recent reports with new flag
        \item GET /api/prescriptions?status=active for active prescriptions with refill info
        \item GET /api/messages/unread/count for notification badge count
    \end{itemize}
    \item Data is cached client-side with appropriate invalidation strategies (5-minute cache for non-critical data)
    \item Real-time updates via WebSocket for appointment changes and new messages
    \item Lazy loading for sections below the fold to improve initial load time
    \item Error boundaries for each component to prevent total dashboard failure if one API fails
\end{itemize}

\subsection{Doctor Dashboard Module}

The doctor dashboard is designed for clinical workflow efficiency, providing quick access to today's schedule, patient information, and pending tasks. The interface prioritizes information density while maintaining clarity and ease of use, following clinical workflow patterns familiar to healthcare providers.

\textbf{Component Architecture:}

\begin{enumerate}
    \item \textbf{Dashboard Header}:
    \begin{itemize}
        \item Welcome message with doctor name and greeting based on time of day
        \item Current date and time with timezone indication
        \item Quick stats: today's patient count (completed/remaining), pending reports count, unread messages
        \item Profile quick-access dropdown with settings, schedule, logout
        \item Status toggle (Available/Busy/Offline) affecting appointment availability display
    \end{itemize}
    
    \item \textbf{Today's Schedule Component}:
    \begin{itemize}
        \item Chronological list of today's appointments with clear time slots
        \item Each appointment card shows comprehensive information:
        \begin{itemize}
            \item Time slot with duration and color-coded status (on time, delayed, completed)
            \item Patient name, age, and brief reason for visit (chief complaint)
            \item Appointment type icon (in-person/video) for quick identification
            \item Status indicator (checked-in, waiting, in-progress, completed, no-show)
            \item Wait time since scheduled time for delayed appointments (highlighted in red)
        \end{itemize}
        \item Action buttons per appointment in consistent layout:
        \begin{itemize}
            \item Start Consultation (opens patient record in modal with all clinical tools)
            \item View History (quick popup of previous visits with diagnoses and prescriptions)
            \item Message Patient (opens chat sidebar with patient context)
            \item Mark No-Show (with reason selection and automatic follow-up)
        \end{itemize}
        \item Color coding for delayed appointments (yellow after 5 min, red after 15 min)
        \item Drag-and-drop for manual rescheduling within same day (with confirmation)
    \end{itemize}
    
    \item \textbf{Patient List Sidebar}:
    \begin{itemize}
        \item Searchable list of all patients under doctor's care with instant search
        \item Quick filters: recent (last 7 days), frequent (top 10), new (first visit)
        \item Each entry shows patient name, last visit date, next appointment (if any)
        \item Hover preview of recent activity (last diagnosis, pending tasks)
        \item Click to open full patient record in main view with context preservation
    \end{itemize}
    
    \item \textbf{Lab Reports Queue}:
    \begin{itemize}
        \item Priority-based organization with clear visual hierarchy:
        \begin{itemize}
            \item \textbf{Critical (red)}: Requires immediate attention - results outside critical thresholds
            \item \textbf{Urgent (orange)}: Review within 24 hours - abnormal but not critical
            \item \textbf{Normal (blue)}: Routine review - within normal ranges
        \end{itemize}
        \item Each report card shows key information:
        \begin{itemize}
            \item Patient name and age for identification
            \item Test type and collection date
            \item Key results with abnormal values highlighted and flagged
            \item Time since report availability with aging indicator
        \end{itemize}
        \item One-click action to add findings (opens findings entry modal with template)
        \item Bulk actions for normal reports (mark all as reviewed)
    \end{itemize}
    
    \item \textbf{Prescription Writing Interface}:
    \begin{itemize}
        \item Medicine search with auto-complete from hospital formulary (including alternatives)
        \item Common dosage templates for quick selection (e.g., ``500mg twice daily'')
        \item Frequency picker with common options (daily, twice daily, every 8 hours, as needed)
        \item Duration calculator with start/end dates and refill count
        \item Drug interaction checker integration (real-time checking as medicines added)
        \item Allergy alert system (warns if prescribed medicine conflicts with patient allergies)
        \item Digital signature capture with timestamp for legal validity
        \item Print/email prescription options with pharmacy selection
    \end{itemize}
    
    \item \textbf{Leave Management Widget}:
    \begin{itemize}
        \item Upcoming leave summary with dates and approval status
        \item Apply for leave button (opens form with date picker and reason dropdown)
        \item Leave balance display (if tracked by hospital policy)
        \item Pending requests status with expected approval timeline
    \end{itemize}
    
    \item \textbf{Quick Statistics Cards}:
    \begin{itemize}
        \item Total patients seen today with comparison to average
        \item Average consultation time with trend indicator
        \item Pending reports count with priority breakdown
        \item Unread messages count with preview of latest
        \item Patient satisfaction rating (if feedback available)
    \end{itemize}
\end{enumerate}

\textbf{Workflow Integration:}
\begin{itemize}
    \item Clicking ``Start Consultation'' opens patient record in a modal with contextual tabs:
    \begin{itemize}
        \item \textbf{Patient Info}: Demographics, conditions, allergies, emergency contacts with visual alerts
        \item \textbf{Medical History}: Past consultations, diagnoses, procedures in timeline view
        \item \textbf{Lab Results}: All reports organized by date with trend visualization for repeated tests
        \item \textbf{Prescriptions}: Active and past medications with adherence tracking if available
        \item \textbf{Notes}: Consultation notes entry area with templates and voice-to-text option
        \item \textbf{Billing}: Payment status and history, insurance information
    \end{itemize}
    \item Real-time updates for new lab reports and messages via WebSocket (toast notifications)
    \item Auto-refresh of schedule every 5 minutes (configurable) with visual indicator of refresh
    \item Keyboard shortcuts for power users (Ctrl+N for new prescription, Ctrl+S for save)
\end{itemize}

\subsection{Admin Dashboard Module}

The admin dashboard provides hospital administrators with comprehensive tools for managing the entire healthcare facility operations. It consolidates data from all modules and presents actionable insights through visual analytics and management interfaces, enabling data-driven decision making and operational efficiency.

\textbf{Component Architecture:}

\begin{enumerate}
    \item \textbf{Statistics Cards Row}:
    \begin{itemize}
        \item \textbf{Total Doctors}: Active/inactive breakdown with click-through to doctor list
        \item \textbf{Today's Appointments}: Count with comparison to yesterday (percentage change)
        \item \textbf{Pending Approvals}: Leave requests, doctor registrations, pending verifications
        \item \textbf{Inventory Status}: Low stock count, expiring medicines (next 30 days)
        \item \textbf{Revenue Today/Month}: Financial snapshot with currency formatting
        \item \textbf{Patient Registrations}: New patients today/this week with trend
        \item \textbf{Lab Tests Pending}: Count by priority (normal/urgent/critical)
    \end{itemize}
    Each card includes trend indicators (up/down arrows) and sparkline charts for quick visualization of historical patterns.
    
    \item \textbf{Quick Action Cards Grid}:
    \begin{itemize}
        \item Add New Doctor (opens multi-step registration form with credential upload)
        \item Add Medicine to Inventory (opens form with barcode scanning option)
        \item Add Lab Technician (opens registration form with role selection)
        \item View All Appointments (opens appointment overview with filters)
        \item Generate Reports (opens report builder with template selection)
        \item Manage Leave Requests (opens leave management view with calendar)
        \item View Financial Summary (opens financial dashboard with charts)
        \item System Settings (opens configuration panel with tabs)
    \end{itemize}
    Each card includes icon, description, and one-click access with keyboard shortcuts.
    
    \item \textbf{Doctor Management Interface}:
    \begin{itemize}
        \item Searchable/filterable table of all doctors with server-side pagination
        \item Columns: Name, Specialization, Department, Status (active/on leave/inactive), Patient Count, Rating, Actions
        \item Inline editing for quick updates (status, department, fees)
        \item Bulk actions for status changes (activate/deactivate selected)
        \item Credential verification workflow with document upload and expiry tracking
        \item Document upload for licenses and certifications with expiration alerts
        \item Schedule management view (weekly calendar with working hours)
        \item Performance metrics integration (patients seen, ratings, revenue)
    \end{itemize}
    
    \item \textbf{Medicine Inventory Management}:
    \begin{itemize}
        \item Real-time stock levels with visual indicators (green: adequate, yellow: low, red: critical)
        \item Low stock alerts with reorder suggestions based on usage patterns
        \item Expiry tracking with color-coded warnings (yellow: 3 months, red: 1 month)
        \item Batch/lot tracking for recall management
        \item Price management with retail and insurance rates
        \item Supplier information and purchase order history
        \item Usage analytics and forecasting for procurement planning
        \item Inventory adjustment with audit trail (reason recorded)
    \end{itemize}
    
    \item \textbf{Lab Technician Management}:
    \begin{itemize}
        \item List of all lab technicians with roles and certifications
        \item Workload distribution view (tests assigned vs. completed)
        \item Performance metrics (tests completed, turnaround time, accuracy)
        \item Schedule management with shift assignments
        \item Certification tracking with expiry alerts
    \end{itemize}
    
    \item \textbf{Leave Approval Workflow}:
    \begin{itemize}
        \item Calendar view showing all leave requests with color coding (pending, approved, rejected)
        \item Department-wise filtering for focused view
        \item Pending requests queue with quick approve/reject buttons and comment field
        \item Coverage analysis showing gaps (days with insufficient coverage)
        \item Leave balance tracking per doctor (accrued, used, remaining)
        \item Automated conflict detection (multiple doctors from same department on leave)
    \end{itemize}
    
    \item \textbf{Appointment Overview}:
    \begin{itemize}
        \item Daily/weekly/monthly appointment volumes with bar charts
        \item Utilization rates by doctor/department (percentage of slots filled)
        \item No-show rates and trends with patient history
        \item Cancellation analysis by reason and time before appointment
        \item Wait time metrics (average, 95th percentile) by doctor
        \item Resource allocation suggestions based on demand patterns
    \end{itemize}
    
    \item \textbf{Analytics and Reporting}:
    \begin{itemize}
        \item Revenue reports with drill-down (by department, doctor, service type)
        \item Patient demographics analysis (age, gender, location)
        \item Popular services and treatments (frequency analysis)
        \item Doctor performance comparisons (patients/hour, satisfaction, revenue)
        \item Custom report builder with drag-and-drop fields and filters
        \item Export capabilities (PDF, Excel, CSV) with formatting options
        \item Scheduled report delivery via email (daily/weekly/monthly)
    \end{itemize}
    
    \item \textbf{System Configuration}:
    \begin{itemize}
        \item Role and permission management (create roles, assign permissions)
        \item Notification template editor (email, SMS, push templates with variables)
        \item Integration settings (API keys, webhook URLs for third-party services)
        \item Audit log viewer with search and export (user actions, timestamps, IPs)
        \item Backup management (schedule backups, test restoration)
        \item System health monitoring (CPU, memory, database performance)
    \end{itemize}
\end{enumerate}

\textbf{Data Visualization:}
\begin{itemize}
    \item Charts for trends and comparisons using Chart.js or similar (line charts for trends, bar for comparisons)
    \item Heat maps for busy periods (appointment density by hour/day)
    \item Geographic maps for patient distribution (cluster view)
    \item Gantt charts for leave planning and resource allocation
    \item Real-time dashboard refresh (every 30 seconds for critical metrics) with WebSocket
\end{itemize}

\subsection{Lab Technician Module}

The laboratory technician module streamlines the complete workflow of test processing, from patient identification through result entry to report submission and notification. It is designed for efficiency and accuracy in high-volume laboratory environments, with validation at every step to ensure data quality and patient safety.

\textbf{Component Architecture:}

\begin{enumerate}
    \item \textbf{Send Report Form}:
    \begin{itemize}
        \item Multi-step form with validation at each step and progress indicator:
        \begin{enumerate}
            \item \textbf{Patient Selection}:
            \begin{itemize}
                \item Search by patient ID, name, or contact with real-time suggestions
                \item Auto-suggest based on recent patients (last 7 days) for speed
                \item QR code scanner integration for mobile/tablet (camera-based)
                \item Patient verification (DOB confirmation, name confirmation)
                \item Display of patient demographics for visual confirmation
            \end{itemize}
            
            \item \textbf{Test Information}:
            \begin{itemize}
                \item Test type dropdown categorized by department (hematology, chemistry, microbiology)
                \item Dynamic form fields based on test type (different inputs for different tests)
                \item Reference ranges display for context (age/gender-specific)
                \item Test date/time picker (defaults to current time, but adjustable)
                \item Ordering doctor selection with search (recent doctors first)
                \item Additional notes field for special instructions
            \end{itemize}
            
            \item \textbf{Result Entry}:
            \begin{itemize}
                \item Numeric fields with unit selection from predefined options
                \item Real-time validation against reference ranges with visual feedback
                \item Color indicators for abnormal values (yellow for abnormal, red for critical)
                \item Text areas for qualitative results with spell check
                \item Multiple entry for panels (multiple results per test with tab order)
                \item Critical value confirmation dialog requiring acknowledgment
                \item Comments and interpretations field for additional context
            \end{itemize}
            
            \item \textbf{File Upload}:
            \begin{itemize}
                \item Drag-and-drop file upload area with visual feedback
                \item Support for PDF, JPG, PNG, TIFF (max 25MB per file)
                \item Multiple file upload capability with progress indicators
                \item File preview before submission (thumbnail for images, first page for PDF)
                \item Automatic file naming convention based on patient and test
                \item Virus scanning status indicator (scanning, clean, infected)
            \end{itemize}
            
            \item \textbf{Report Configuration}:
            \begin{itemize}
                \item Priority selection radio buttons (normal, urgent, critical)
                \item Priority impact explanation tooltips (notification differences)
                \item Sharing preferences (doctor only, patient only, both) with radio buttons
                \item Additional recipients option (email addresses for copy)
                \item Confidentiality notice acknowledgment checkbox
                \item Technician signature/ID field (auto-filled from login)
            \end{itemize}
            
            \item \textbf{Review and Submit}:
            \begin{itemize}
                \item Summary of all entered data in organized sections
                \item Edit buttons for each section to go back and correct
                \item Validation summary (all required fields completed)
                \item Submit button with confirmation dialog
                \item Save as draft option for partial completion
            \end{itemize}
        \end{enumerate}
    \end{itemize}
    
    \item \textbf{Patient History Table}:
    \begin{itemize}
        \item Searchable table of patients with recent tests (last 30 days)
        \item Columns: Patient Name, ID, Recent Tests (count), Last Test Date, Actions
        \item Click to view complete test history with trend graphs for repeated tests
        \item Trend visualization for repeated tests (line charts for quantitative results)
        \item Previous results for comparison (displayed alongside new results)
        \item Flag for patients with pending reports (needs attention)
    \end{itemize}
    
    \item \textbf{Statistics Dashboard}:
    \begin{itemize}
        \item \textbf{Today's Tests}: Total completed today with comparison to yesterday
        \item \textbf{Pending Reports}: Count awaiting processing (by priority)
        \item \textbf{Completed Today}: Successfully submitted reports with breakdown
        \item \textbf{Urgent Cases}: High-priority pending tests requiring immediate attention
        \item \textbf{Average Turnaround Time}: Performance metric (time from order to completion)
        \item \textbf{Your Queue}: Personal workload (assigned tests)
        \item \textbf{Department Statistics}: Team performance with averages
    \end{itemize}
    
    \item \textbf{Quality Control Panel}:
    \begin{itemize}
        \item Control results entry for quality assurance samples
        \item QC charts and trending (Levey-Jennings plots)
        \item Equipment status tracking (calibration due, maintenance)
        \item Reagent lot tracking with expiry alerts
        \item Proficiency testing records and results
    \end{itemize}
    
    \item \textbf{Recent Activity Feed}:
    \begin{itemize}
        \item Chronological list of recent submissions (last 24 hours)
        \item Status updates (submitted, reviewed by doctor, printed)
        \item Quick reprint/re-send options for troubleshooting
        \item Error notifications and resolutions (failed uploads, validation errors)
    \end{itemize}
\end{enumerate}

\textbf{Workflow Optimization:}
\begin{itemize}
    \item Keyboard shortcuts for common actions (Ctrl+N new report, Ctrl+S save, Ctrl+F search)
    \item Default values for frequent tests (pre-filled based on last entry)
    \item Templates for common result patterns (e.g., ``Normal'' for routine results)
    \item Auto-save of drafts every 30 seconds to prevent data loss
    \item Batch processing for multiple samples from same patient (group entry)
    \item Integration with laboratory equipment for direct data import (LIS interface)
    \item Barcode scanning for sample tracking and error reduction
\end{itemize}

\subsection{Appointment Booking Module}

The appointment booking module provides patients with an intuitive interface for finding and scheduling appointments with doctors. It incorporates real-time availability, intelligent suggestions, and comprehensive booking management features to ensure a seamless experience.

\textbf{Component Architecture:}

\begin{enumerate}
    \item \textbf{Doctor Search Interface}:
    \begin{itemize}
        \item Search bar for doctor name, specialty, hospital with autocomplete
        \item Advanced filters expandable panel:
        \begin{itemize}
            \item Specialization (multi-select with department grouping)
            \item Hospital/Location (with distance sorting if location enabled)
            \item Available dates (calendar picker with availability indicators)
            \item Gender preference (male, female, any)
            \item Languages spoken (multi-select)
            \item Insurance accepted (search by insurance provider)
            \item Rating range (slider for minimum rating, e.g., 4+ stars)
            \item Consultation fee range (slider with currency)
            \item Available today/tomorrow/this week (quick filters)
        \end{itemize}
        \item Map view for location-based search (shows doctors near user)
        \item Sort options (by rating, experience, fee low-high, next available)
        \item Search results display with clear card layout:
        \begin{itemize}
            \item Doctor photo, name, specialization for quick identification
            \item Hospital name and distance (if location available)
            \item Rating stars and review count with breakdown
            \item Consultation fee with insurance indicator
            \item Next available slot (e.g., ``Tomorrow 10:30 AM'')
            \item Book button with availability preview
        \end{itemize}
    \end{itemize}
    
    \item \textbf{Doctor Profile View}:
    \begin{itemize}
        \item Detailed information organized in tabs:
        \begin{itemize}
            \item \textbf{About}: Complete qualifications, certifications, and bio
            \item \textbf{Experience}: Years of experience, previous positions, special training
            \item \textbf{Specializations}: Areas of expertise with descriptions
            \item \textbf{Languages}: Languages spoken with fluency levels
            \item \textbf{Fees}: Consultation fees (in-person/video) with payment options
            \item \textbf{Location}: Complete address with map, parking information, photos of facility
            \item \textbf{Reviews}: Patient reviews and ratings with date and verified badge
            \item \textbf{Availability}: Calendar view with available slots
        \end{itemize}
    \end{itemize}
    
    \item \textbf{Availability Calendar}:
    \begin{itemize}
        \item Monthly/weekly/day view options with toggle
        \item Color-coded availability:
        \begin{itemize}
            \item Green: Available slots (multiple)
            \item Yellow: Limited availability (1-2 slots)
            \item Red: Fully booked
            \item Gray: Doctor not working (off day, leave)
        \end{itemize}
        \item Click on date to see detailed time slots below calendar
        \item Quick navigation to next available date (one-click)
        \item Timezone awareness (converts to user's timezone)
        \item Buffer times between appointments shown as blocked
        \item Holiday and leave indicators with tooltips
    \end{itemize}
    
    \item \textbf{Time Slot Selection}:
    \begin{itemize}
        \item Grid of available time slots for selected date (15/30/60-minute increments)
        \item Slots shown with time and duration (e.g., ``10:30 AM (30 min)'')
        \item Real-time availability with auto-refresh every 30 seconds
        \item Slot locking during selection (5-minute hold while completing booking)
        \item Duration indication for different appointment types (consultation, procedure)
        \item Preferred slot highlighting based on user history
        \item Waitlist option for fully booked slots (notify when available)
    \end{itemize}
    
    \item \textbf{Booking Form}:
    \begin{itemize}
        \item Reason for visit (text area with common templates: ``New consultation'', ``Follow-up'', ``Urgent concern'')
        \item Symptoms description with structured input (onset, duration, severity)
        \item Previous visit reference (if follow-up, select from history)
        \item Appointment type selector (in-person/video) with availability check
        \item Insurance information selection (from saved profiles)
        \item Special requests or notes for doctor
        \item Emergency contact confirmation (verify or update)
    \end{itemize}
    
    \item \textbf{Confirmation Modal}:
    \begin{itemize}
        \item Appointment summary with clear formatting:
        \begin{itemize}
            \item Doctor name and photo for verification
            \item Date and time with timezone
            \item Location with map link or video meeting link
            \item Fee breakdown (consultation, taxes, total)
            \item Token number (if applicable) for physical visits
            \item Cancellation policy reminder (time window, fees)
        \end{itemize}
        \item Calendar integration buttons (Google, Apple, Outlook) with one-click
        \item Add to phone calendar option with .ics download
        \item Set reminder preferences (when, how)
        \item Proceed to payment button (if required) with amount
        \item Confirm and book button (final step)
    \end{itemize}
    
    \item \textbf{Rescheduling Interface}:
    \begin{itemize}
        \item Similar to booking flow but with original appointment context
        \item Original appointment details displayed for reference
        \item Reason for rescheduling required (dropdown + optional text)
        \item Availability shown with current appointment time blocked
        \item Notification preview (who will be notified)
        \item Rescheduling fee indicator (if applicable based on policy)
    \end{itemize}
    
    \item \textbf{Cancellation Interface}:
    \begin{itemize}
        \item Reason selection (dropdown: changed mind, found alternative, emergency, other)
        \item Refund eligibility calculator based on cancellation policy
        \item Confirmation with policy reminder and consequences
        \item Feedback request option (optional survey)
        \item Notification preview (who will be notified)
    \end{itemize}
\end{enumerate}

\textbf{Business Logic:}
\begin{itemize}
    \item Slot availability calculated as: (doctor working hours - booked appointments - leave periods - buffer times)
    \item Concurrent booking prevention through database transactions with row-level locking
    \item Waitlist management with auto-booking when slots open (first-come, first-served)
    \item Intelligent slot suggestions based on patient history (preferred times)
    \item Machine learning for appointment duration prediction based on reason and patient history
\end{itemize}

\subsection{Payment Integration Module}

The payment integration module handles all financial transactions securely, integrating with multiple payment gateways while maintaining comprehensive internal records for reconciliation and audit purposes. It is designed to support multiple payment methods, currencies, and regional requirements.

\textbf{Component Architecture:}

\begin{enumerate}
    \item \textbf{Payment Gateway Abstraction Layer}:
    \begin{itemize}
        \item Unified interface for multiple gateways (Stripe, Razorpay, PayPal)
        \item Gateway-specific adapters implementing common interface (create payment, refund, webhook)
        \item Automatic fallback to secondary gateway on primary failure (circuit breaker pattern)
        \item Webhook handling with signature verification for each gateway
        \item Idempotency keys to prevent duplicate processing (retry-safe)
    \end{itemize}
    
    \item \textbf{Payment Intent Creation}:
    \begin{itemize}
        \item Creates payment intent with gateway via API
        \item Returns client secret for front-end tokenization (PCI compliance)
        \item Stores intent ID with payment record for reconciliation
        \item Sets expiration time for payment (30 minutes typical)
        \item Includes metadata (appointment ID, patient ID) for webhook processing
    \end{itemize}
    
    \item \textbf{Payment Method Selection UI}:
    \begin{itemize}
        \item Card payment with PCI-compliant iframe (Stripe Elements)
        \item Saved cards management (tokenized, last 4 digits display)
        \item Digital wallets (Apple Pay, Google Pay) with device detection
        \item Insurance billing information capture (policy number, group ID)
        \item Cash payment recording (for admin use in clinics)
        \item UPI/Bank transfer options (region-specific, QR code display)
    \end{itemize}
    
    \item \textbf{Payment Processing Flow}:
    \begin{itemize}
        \item Amount display with clear breakdown:
        \begin{itemize}
            \item Consultation fee (base amount)
            \item Taxes (GST/VAT percentage and amount)
            \item Discounts applied (promotional, insurance)
            \item Total amount due
        \end{itemize}
        \item Loading states during processing with spinner and timeout indicator
        \item Success/failure/pending screens with appropriate messaging
        \item Retry logic for failed payments (3 attempts with exponential backoff)
        \item Timeout handling (30 seconds, then show timeout message)
    \end{itemize}
    
    \item \textbf{Invoice Generation Engine}:
    \begin{itemize}
        \item PDF generation using templates (jsPDF or server-side library)
        \item Includes comprehensive information:
        \begin{itemize}
            \item Hospital logo and details (name, address, tax ID)
            \item Patient name and ID
            \item Appointment details (doctor, date, time)
            \item Itemized charges with descriptions
            \item Tax breakdown by rate
            \item Payment reference and transaction ID
            \item QR code for verification (scan to validate)
        \end{itemize}
        \item Email delivery with attachment (PDF)
        \item Storage in cloud with access URL (signed URL with expiry)
        \item Invoice numbering sequence (unique, sequential per hospital)
        \item GST/Tax invoice compliance for regions requiring it
    \end{itemize}
    
    \item \textbf{Refund Processing Module}:
    \begin{itemize}
        \item Refund eligibility check based on cancellation policy
        \item Proration calculation for partial refunds (percentage or fixed)
        \item Gateway refund API integration with appropriate parameters
        \item Approval workflow for large refunds (>threshold requires second admin)
        \item Partial refund support (amount field)
        \item Refund reason tracking (dropdown + text)
        \item Credit note generation (separate document)
    \end{itemize}
    
    \item \textbf{Payment History and Reconciliation}:
    \begin{itemize}
        \item Transaction listing with filters (date range, status, method)
        \item Settlement tracking (pending, settled, failed)
        \item Dispute management (view, respond, track status)
        \item Reconciliation reports (transactions vs. gateway statements)
        \item Export for accounting systems (CSV, Excel, QuickBooks format)
        \item Audit log of all financial actions (who, what, when)
    \end{itemize}
    
    \item \textbf{Webhook Handler}:
    \begin{itemize}
        \item Secure endpoint with signature verification (gateway-specific)
        \item Event type processing (payment.succeeded, payment.failed, refund.completed)
        \item Idempotent processing via event IDs (ignore duplicates)
        \item Automatic retry on failure with exponential backoff
        \item Logging for debugging and audit (request/response)
    \end{itemize}
\end{enumerate}

\textbf{Security Measures:}
\begin{itemize}
    \item PCI DSS compliance through tokenization (no storage of raw card details)
    \item Encryption of sensitive data at rest (AES-256 for stored financial data)
    \item HTTPS for all communications (TLS 1.3)
    \item Fraud detection rules (velocity checks, amount thresholds, geographic anomalies)
    \item 3D Secure support for additional authentication
    \item Rate limiting on payment endpoints (5 per minute per user)
\end{itemize}

\subsection{Real-time Chat Module}

The real-time chat module enables secure, instant communication between patients and doctors, supporting text messages, file sharing, and read receipts. Built on Socket.io, it provides a responsive messaging experience while maintaining message persistence for medical record keeping and legal compliance.

\textbf{Component Architecture:}

\begin{enumerate}
    \item \textbf{WebSocket Server} (Socket.io):
    \begin{itemize}
        \item Handles real-time bidirectional communication with low latency
        \item Authentication middleware for connection establishment (JWT validation)
        \item Room-based architecture for private conversations (one room per conversation)
        \item Scalability through Redis adapter for multi-server deployments
        \item Connection state management and reconnection handling (automatic reconnect)
        \item Heartbeat mechanism for connection health monitoring (ping/pong)
    \end{itemize}
    
    \item \textbf{Message Persistence Layer}:
    \begin{itemize}
        \item All messages stored in PostgreSQL messages table with indexes
        \item Indexed for fast conversation retrieval (sender\_id, receiver\_id, created\_at)
        \item Encryption at rest for sensitive medical discussions
        \item Retention policy compliance with healthcare regulations (7 years)
        \item Archive mechanism for old conversations (move to cold storage)
    \end{itemize}
    
    \item \textbf{Chat List Interface}:
    \begin{itemize}
        \item Displays all recent conversations for the user with previews
        \item Each conversation item shows:
        \begin{itemize}
            \item Other participant's name and photo (avatar)
            \item Last message preview (truncated to 50 characters)
            \item Timestamp of last message (relative, e.g., ``5 min ago'')
            \item Unread message count badge (red circle with number)
            \item Online/offline status indicator (green dot for online)
            \item Typing indicator when applicable (animated dots)
        \end{itemize}
        \item Search functionality to find specific conversations (by participant name)
        \item Pin important conversations to top (star icon)
        \item Archive old conversations (hide from main list)
        \item Delete conversation option (with confirmation and audit)
    \end{itemize}
    
    \item \textbf{Message Thread View}:
    \begin{itemize}
        \item Chronological display of messages in a conversation with infinite scroll
        \item Date separators for messages from different days (e.g., ``Yesterday'', ``March 4'')
        \item Message bubbles with clear sender identification:
        \begin{itemize}
            \item Sent messages: Right-aligned, colored background (blue)
            \item Received messages: Left-aligned, light background (gray)
            \item System messages: Centered, grey (e.g., ``Patient joined the chat'')
        \end{itemize}
        \item Read receipts indicators (single tick: sent, double tick: delivered, blue double tick: read)
        \item Delivery status for offline messages (sent when user comes online)
        \item Timestamp display on hover or for each message (actual time)
        \item File attachment preview (images shown inline, documents as icons with filename)
        \item Download button for attachments with preview option
    \end{itemize}
    
    \item \textbf{Message Input Area}:
    \begin{itemize}
        \item Rich text input with emoji picker (Emoji Mart integration)
        \item File attachment button with type/size validation (max 25MB)
        \item Supported file types: Images (JPG, PNG, GIF), Documents (PDF, DOC, DOCX)
        \item Image preview before sending (crop, rotate options)
        \item Voice message recording (optional integration with Web Audio API)
        \item Typing indicator broadcast (shows ``typing...'' to other participant)
        \item Send on Enter (configurable, Ctrl+Enter for new line)
        \item Character counter for long messages (max 2000 characters)
    \end{itemize}
    
    \item \textbf{Push Notification Service}:
    \begin{itemize}
        \item Real-time notifications for new messages when user is offline
        \item Integration with Firebase Cloud Messaging (FCM) for mobile apps
        \item Web Push API for desktop browsers (service workers)
        \item Notification content customization (show/hide message preview based on settings)
        \item Sound alerts with mute option per conversation
        \item Do Not Disturb hours configuration (global and per user)
    \end{itemize}
    
    \item \textbf{File Sharing Service}:
    \begin{itemize}
        \item Secure file upload to cloud storage (AWS S3/Azure Blob) with unique naming
        \item Virus scanning before storage (ClamAV integration)
        \item File type validation by content (not just extension)
        \item Size limit enforcement (25MB per file)
        \item Automatic image compression for large files (resize to 1080p)
        \item Expiring URLs for secure access (5-minute validity)
        \item File encryption at rest (AES-256)
    \end{itemize}
    
    \item \textbf{Conversation History}:
    \begin{itemize}
        \item Infinite scroll with pagination (load 20 messages at a time)
        \item Search within conversation (full-text search)
        \item Export conversation as PDF (for medical records)
        \item Print conversation option (print-friendly formatting)
        \item Date-based navigation (jump to specific date)
    \end{itemize}
\end{enumerate}

\textbf{WebSocket Events:}

\begin{itemize}
    \item \texttt{connection}: Client connects to WebSocket server
    \item \texttt{authenticate}: Client sends JWT for authentication
    \item \texttt{join\_conversation}: Client joins a specific conversation room (based on conversation ID)
    \item \texttt{leave\_conversation}: Client leaves conversation room
    \item \texttt{send\_message}: Client sends a new message (content, attachment reference)
    \item \texttt{message\_received}: Server acknowledges message receipt (confirmation)
    \item \texttt{new\_message}: Server broadcasts new message to recipient (real-time)
    \item \texttt{typing\_start}: Client indicates typing started
    \item \texttt{typing\_stop}: Client indicates typing stopped
    \item \texttt{message\_read}: Client marks message as read
    \item \texttt{read\_receipt}: Server broadcasts read receipt to sender
    \item \texttt{error}: Server sends error information (with code)
    \item \texttt{disconnect}: Client disconnects from server (cleanup)
\end{itemize}

\textbf{Security Features:}
\begin{itemize}
    \item End-to-end encryption for sensitive medical discussions (optional, using Web Crypto API)
    \item Authentication required for all WebSocket connections (JWT validation)
    \item Rate limiting to prevent spam (10 messages per minute per user)
    \item Message filtering for prohibited content (PHI, profanity)
    \item Audit logging of all communications (who, when, what)
    \item Automatic logout on session expiry (WebSocket disconnect)
    \item HIPAA compliance through encryption and access controls
\end{itemize}

\subsection{Notification Module}

The notification module ensures timely communication across all platform events, delivering messages through multiple channels based on user preferences and event urgency. It handles appointment reminders, lab report notifications, prescription alerts, and system announcements with reliability and personalization.

\textbf{Component Architecture:}

\begin{enumerate}
    \item \textbf{Notification Types and Templates}:
    \begin{itemize}
        \item \textbf{Appointment Notifications}:
        \begin{itemize}
            \item Confirmation: Sent immediately after booking (with details)
            \item Reminder: 24 hours before appointment (with preparation instructions)
            \item Reminder: 2 hours before appointment (for same-day, with check-in link)
            \item Rescheduling: Notifies all parties of change (with new details)
            \item Cancellation: Notifies all affected parties (with reason if available)
            \item Follow-up: After appointment completion (with feedback request)
        \end{itemize}
        
        \item \textbf{Lab Report Notifications}:
        \begin{itemize}
            \item Report Ready: Alert to patient and/or doctor based on sharing preferences
            \item Critical Result: Immediate alert with escalation (SMS + phone call)
            \item Report Reviewed: Doctor has reviewed results (patient notified)
            \item Report Updated: When corrections are made (with version info)
        \end{itemize}
        
        \item \textbf{Prescription Notifications}:
        \begin{itemize}
            \item New Prescription: When doctor issues prescription (with medicine list)
            \item Refill Available: When refill is due (automatic reminder)
            \item Refill Request: Patient requests refill (doctor notified)
            \item Refill Approved/Rejected: Status update to patient (with reason)
            \item Expiring Soon: Medication about to expire (with refill prompt)
        \end{itemize}
        
        \item \textbf{Medicine Reminders}:
        \begin{itemize}
            \item Daily Medication: Scheduled dosage reminders (with medicine name)
            \item Refill Reminder: Low stock alert (reorder suggestion)
            \item Missed Dose: Follow-up if dose not marked taken (with snooze)
        \end{itemize}
        
        \item \textbf{System Notifications}:
        \begin{itemize}
            \item Account Creation: Welcome email with verification link
            \item Password Reset: Security notifications with token
            \item Profile Updates: Confirmation of changes
            \item Security Alerts: Unusual activity detected (new device, location)
        \end{itemize}
    \end{itemize}
    
    \item \textbf{Multi-Channel Delivery Engine}:
    \begin{itemize}
        \item \textbf{In-App Notifications}:
        \begin{itemize}
            \item Real-time display in notification center (dropdown)
            \item Badge counters on navigation icons (unread count)
            \item Toast/popup notifications for urgent messages (auto-dismiss)
            \item Mark as read/unread functionality (individual or all)
            \item Clear all option for bulk actions
            \item Persistent storage of notification history (30 days)
        \end{itemize}
        
        \item \textbf{Email Notifications} (via SendGrid/NodeMailer):
        \begin{itemize}
            \item HTML email templates with responsive design (mobile-friendly)
            \item Plain text alternative for compatibility (email clients)
            \item Personalization with user name and dynamic content
            \item Tracking pixels for open rates (with privacy opt-out)
            \item Unsubscribe links for non-transactional notifications
            \item DKIM/SPF compliance for deliverability
        \end{itemize}
        
        \item \textbf{SMS Notifications} (via Twilio/MessageBird):
        \begin{itemize}
            \item Concise messages with essential information (160 chars optimized)
            \item Short links to relevant pages (bit.ly integration)
            \item Opt-out support (STOP keyword for compliance)
            \item Delivery status tracking (delivered, failed)
            \item Rate limiting to prevent spam (per number)
            \item Cost optimization through message bundling
        \end{itemize}
        
        \item \textbf{Push Notifications} (via Firebase/Web Push):
        \begin{itemize}
            \item Mobile app push notifications (iOS, Android)
            \item Web browser push notifications (Chrome, Firefox, Safari)
            \item Action buttons in notifications (e.g., ``Confirm'', ``View'', ``Snooze'')
            \item Silent notifications for data sync (background updates)
            \item Priority levels for delivery (high for critical alerts)
        \end{itemize}
    \end{itemize}
    
    \item \textbf{Template Management System}:
    \begin{itemize}
        \item Centralized template storage in database (versioned)
        \item Variable substitution using \{\{variable\}\} syntax (e.g., \{\{patient\_name\}\})
        \item Multi-language support (English, Spanish, Hindi, etc.)
        \item Version control for templates (history, revert)
        \item Preview functionality before sending (test with sample data)
        \item A/B testing capability for marketing notifications
        \item Template inheritance for consistent branding
    \end{itemize}
    
    \item \textbf{Notification Scheduler}:
    \begin{itemize}
        \item Cron jobs for scheduled notifications (reminders)
        \item Queue-based processing for reliability (Bull/Redis)
        \item Retry logic for failed deliveries (exponential backoff, max 3 attempts)
        \item Dead letter queue for persistent failures (manual review)
        \item Timezone awareness for reminders (send at local time)
        \item Throttling to prevent overwhelming users (max 5 per hour)
        \item Batching of non-urgent notifications (daily digest)
    \end{itemize}
    
    \item \textbf{User Preferences Manager}:
    \begin{itemize}
        \item Per-channel opt-in/opt-out (email, SMS, push, in-app)
        \item Quiet hours configuration (e.g., 10 PM - 8 AM no notifications)
        \item Frequency preferences (immediate, daily digest, weekly summary)
        \item Notification type granularity (appointment vs. marketing)
        \item Device-specific settings (mobile vs. desktop)
        \item Preference persistence and sync across devices
    \end{itemize}
    
    \item \textbf{Delivery Analytics}:
    \begin{itemize}
        \item Delivery success/failure tracking by channel
        \item Open rates for email (unique opens)
        \item Click-through rates (links clicked)
        \item Opt-out tracking (unsubscribes by channel)
        \item Channel effectiveness analysis (which channel works best)
        \item Cost per notification reporting (for budget optimization)
    \end{itemize}
\end{enumerate}

\textbf{Queue Architecture:}
\begin{itemize}
    \item Bull/Redis queue for notification jobs with priorities
    \item Priority queues (critical > transactional > promotional)
    \item Delayed jobs for scheduled notifications (future send times)
    \item Failed job retry with exponential backoff (3 attempts)
    \item Job deduplication to prevent duplicate sends (same event within window)
    \item Rate limiting per user/channel (configurable)
    \item Monitoring dashboard for queue health (Bull Board)
\end{itemize}

% =============================================================================
% 4.7 SECURITY ARCHITECTURE
% =============================================================================

\section{Security Architecture}

The BondHealth platform implements a comprehensive, defense-in-depth security architecture to protect sensitive healthcare data, ensure regulatory compliance (HIPAA, GDPR), and maintain system integrity. Security is integrated at every layer of the application, from network perimeter to database storage, with multiple overlapping controls ensuring that no single point of failure compromises patient data.

\textbf{Security Principles:}

\begin{itemize}
    \item \textbf{Defense in Depth}: Multiple layers of security controls ensure that if one layer fails, others continue to provide protection. This includes network security, application security, data security, and physical security layers.
    
    \item \textbf{Least Privilege}: Users and services have only the minimum permissions necessary to perform their functions. No user has access to data they don't need, and service accounts are similarly restricted.
    
    \item \textbf{Secure by Default}: Default configurations are secure, requiring explicit actions to reduce security. Features are off unless needed, and secure options are preselected.
    
    \item \textbf{Fail Secure}: When failures occur, the system fails into a secure state rather than an insecure one. Errors default to denying access rather than granting it.
    
    \item \textbf{Complete Mediation}: Every access attempt is checked for authorization, not just the first. Each request is validated independently.
    
    \item \textbf{Separation of Duties}: Critical operations require multiple people or steps. For example, large refunds require two admin approvals.
    
    \item \textbf{Privacy by Design}: Privacy considerations are built into the system from the ground up, not added as an afterthought.
\end{itemize}

\begin{table}[h]
    \centering
    \caption{Security Features Summary - Comprehensive overview of security controls implemented across all layers of the BondHealth platform. Each control is described with its implementation details and purpose.}
    \label{tab:security-features}
    \begin{tabular}{|p{4cm}|p{5cm}|p{5cm}|}
        \hline
        \textbf{Security Layer} & \textbf{Implementation} & \textbf{Description} \\
        \hline
        Authentication & JWT with HTTP-only cookies & Tokens stored securely in HTTP-only cookies with Secure, HttpOnly, and SameSite flags preventing JavaScript access and CSRF attacks. 7-day expiration with refresh token rotation. Session invalidation on password change. \\
        \hline
        Password Security & Bcrypt with 10 salt rounds & Industry-standard password hashing with automatic salt generation. Minimum 8 characters with complexity requirements (uppercase, lowercase, number, special). Rate limiting on login attempts (5 attempts per 15 minutes) with account lockout. \\
        \hline
        Database Security & SSL connection to Neon PostgreSQL & All database connections require TLS 1.2+ encryption. Connection strings stored in environment variables, never in code. Database access restricted by IP whitelist. Row-level security for multi-tenant isolation. \\
        \hline
        Session Management & Server-side session tracking with Redis & Active sessions tracked in Redis with automatic expiration (30 minutes inactivity). Concurrent session limiting (max 5 per user). Force logout capability from any device. Session invalidation on password change or admin action. \\
        \hline
        Role-Based Access & Middleware route protection with RBAC & Hierarchical role system with permission inheritance. Route-level and data-level access controls. Hospital-scoped data isolation for multi-tenant security (doctors only see their patients). \\
        \hline
        Data Encryption & HTTPS/TLS for all communications & TLS 1.3 for all client-server communications. HSTS headers enforce HTTPS (max-age=31536000). Perfect forward secrecy with modern cipher suites. Certificates from trusted CA, auto-renewed. \\
        \hline
        Input Validation & Server-side validation with sanitization & All input validated against JSON schemas. SQL injection prevention through parameterized queries (no string concatenation). XSS prevention via output encoding (context-aware). HTML sanitization for rich text (DOMPurify). \\
        \hline
        Audit Logging & Database triggers for critical operations & Comprehensive logging of all authentication attempts, data access, and modifications. Tamper-evident logs with cryptographic verification (hash chains). 7-year retention for compliance. Logs centralized in ELK stack. \\
        \hline
        Compliance & HIPAA/GDPR measures & Data anonymization for analytics. Consent management for marketing. Breach notification procedures (72-hour window). Data portability and deletion capabilities (right to be forgotten). Business associate agreements with all vendors. \\
        \hline
        API Security & Rate limiting and input validation & Per-endpoint rate limiting (100-1000 requests per minute based on sensitivity). API key rotation (quarterly). Request size limits (10MB). CORS policy enforcement (whitelisted domains). \\
        \hline
        File Security & Malware scanning and validation & All uploaded files scanned with ClamAV (real-time). File type validation by content magic numbers (not extension). Size limits (25MB). Secure storage with encrypted filenames (UUID). Expiring access URLs (5-minute validity). \\
        \hline
    \end{tabular}
\end{table}

\subsection{Authentication Security}

\textbf{JWT Implementation Details:}
\begin{itemize}
    \item Algorithm: HS256 (HMAC with SHA-256) for symmetric signing
    \item Secret key: 256-bit random value stored in environment variables, rotated quarterly
    \item Payload includes: user\_id, username, role, session\_id, issued\_at (iat), expiration (exp)
    \item Token expiration: 7 days for access tokens (configurable per deployment)
    \item Refresh tokens: 30 days, stored in database with revocation capability
    \item Token versioning: Each user has token version counter, incremented on password change
\end{itemize}

\textbf{Session Management:}
\begin{itemize}
    \item Redis store with key: \texttt{session:\{session\_id\}} and TTL matching token
    \item Session data: user\_id, ip\_address, user\_agent, created\_at, last\_activity
    \item Session timeout: 30 minutes of inactivity (sliding expiration, reset on activity)
    \item Maximum concurrent sessions: 5 per user (configurable, blocks new logins when exceeded)
    \item Session invalidation triggers: password change, admin force logout, account deactivation, suspicious activity
\end{itemize}

\textbf{Password Policy:}
\begin{itemize}
    \item Minimum length: 8 characters
    \item Complexity: at least one uppercase, one lowercase, one number, one special character
    \item Password history: last 5 passwords prevented from reuse (hash comparison)
    \item Maximum age: 90 days (optional, configurable per organization)
    \item Account lockout: after 5 failed attempts (15-minute lockout, increments with each attempt)
    \item Password reset: email with expiring token (24-hour validity, single use)
\end{itemize}

\subsection{Data Security}

\textbf{Encryption at Rest:}
\begin{itemize}
    \item Database: Transparent Data Encryption (TDE) at storage level (AES-256)
    \item Sensitive fields (PII, medical data): Application-level encryption with AES-256-GCM
    \item File storage: Server-side encryption with AWS KMS/Azure Key Vault (AES-256)
    \item Backups: Encrypted with separate keys (offsite storage)
    \item Key management: Hardware Security Module (HSM) for master keys, key rotation quarterly
\end{itemize}

\textbf{Encryption in Transit:}
\begin{itemize}
    \item TLS 1.3 minimum for all external communications (no older protocols)
    \item Internal service communications: mTLS for service mesh (Istio/Linkerd)
    \item Database connections: TLS with certificate validation (mutual TLS)
    \item API endpoints: HSTS preload, perfect forward secrecy (ECDHE)
\end{itemize}

\textbf{Data Classification:}
\begin{itemize}
    \item \textbf{Restricted} (PHI/PII): Medical records, personal information - encryption required, access logging, minimal retention (7 years)
    \item \textbf{Confidential}: Financial data, appointment details - encryption required, access controls, audit logging
    \item \textbf{Internal}: Operational data - standard protection, access controls
    \item \textbf{Public}: Hospital information, doctor profiles - no restrictions, cacheable
\end{itemize}

\subsection{Application Security}

\textbf{Input Validation:}
\begin{itemize}
    \item Schema validation using Joi/express-validator (strict typing)
    \item Type checking and coercion (numbers, booleans, dates)
    \item Range validation for numeric fields (min/max)
    \item Pattern matching for emails (RFC-compliant regex), phone numbers (E.164 format)
    \item Length limits on all string fields (prevent DoS)
    \item Sanitization of HTML content (DOMPurify, remove scripts)
    \item SQL injection prevention via parameterized queries (prepared statements)
\end{itemize}

\textbf{Output Encoding:}
\begin{itemize}
    \item Context-aware encoding (HTML entities, JavaScript strings, CSS, URLs)
    \item Content Security Policy (CSP) headers with strict directives (default-src 'self')
    \item X-Content-Type-Options: nosniff (prevent MIME sniffing)
    \item X-Frame-Options: DENY (prevent clickjacking)
    \item X-XSS-Protection: 1; mode=block (legacy browsers)
\end{itemize}

\textbf{CSRF Protection:}
\begin{itemize}
    \item SameSite=Strict cookies for all session cookies (modern browsers)
    \item CSRF tokens for state-changing operations (double-submit pattern)
    \item Origin and Referer header validation (server-side check)
    \item Idempotent operations for GET requests (no side effects)
\end{itemize}

\subsection{Infrastructure Security}

\textbf{Network Security:}
\begin{itemize}
    \item VPC with private subnets for databases (no public internet access)
    \item Security groups with least privilege access (IP whitelisting)
    \item Web Application Firewall (WAF) for DDoS protection (AWS WAF/Cloudflare)
    \item API Gateway for rate limiting and throttling (1000 requests per minute per IP)
    \item VPN access for administrative functions (AWS Client VPN)
    \item Network segmentation between tiers (web, app, db in separate subnets)
\end{itemize}

\textbf{Server Hardening:}
\begin{itemize}
    \item Minimal base images with only necessary packages (Alpine Linux)
    \item Regular security patches and updates (weekly automated scans)
    \item Host-based intrusion detection (HIDS) with Wazuh/OSSEC
    \item File integrity monitoring (Tripwire/AIDE)
    \item Root login disabled, SSH key authentication only
    \item Principle of least privilege for service accounts (non-root users)
\end{itemize}

\textbf{Container Security:}
\begin{itemize}
    \item Image scanning for vulnerabilities (Trivy/Clair) in CI/CD
    \item Read-only root filesystems (immutable infrastructure)
    \item Non-root user execution (containers run as user 1000)
    \item Resource limits to prevent DoS (CPU/memory limits)
    \item Secrets management (no secrets in images, use HashiCorp Vault)
    \item Runtime security monitoring (Falco for anomalous activity)
\end{itemize}

\subsection{Compliance and Audit}

\textbf{HIPAA Compliance:}
\begin{itemize}
    \item Business Associate Agreements with all vendors (signed)
    \item Access controls and audit trails for all PHI (who accessed what, when)
    \item Encryption of PHI at rest and in transit (AES-256, TLS 1.3)
    \item Breach notification procedures (72-hour notification)
    \item Minimum necessary access principle (role-based)
    \item Automatic logout after inactivity (15 minutes)
    \item Emergency access procedures (break-glass accounts with audit)
\end{itemize}

\textbf{GDPR Compliance:}
\begin{itemize}
    \item Consent management for data processing (explicit opt-in)
    \item Right to access and data portability (export user data)
    \item Right to erasure (right to be forgotten) with verification
    \item Data processing records (Article 30 documentation)
    \item Privacy by design in all features (DPIA for new features)
    \item Data Protection Impact Assessments (DPIA) for high-risk processing
\end{itemize}

\textbf{Audit Logging:}
\begin{itemize}
    \item Log all authentication attempts (success/failure with IP, user agent)
    \item Log all access to PHI (who, what, when, why - patient ID, record type)
    \item Log all data modifications (before/after values for critical fields)
    \item Tamper-evident logging with cryptographic hashes (hash chains)
    \item Centralized log aggregation (ELK stack) with search and alerting
    \item 7-year log retention for compliance (cold storage after 90 days)
    \item Regular log review and anomaly detection (automated alerts)
\end{itemize}

\textbf{Security Monitoring:}
\begin{itemize}
    \item SIEM integration for real-time alerting (Splunk/ELK)
    \item Anomaly detection for unusual access patterns (ML-based)
    \item Brute force attack detection (multiple failures from same IP)
    \item Unusual geographic access alerts (impossible travel)
    \item Data exfiltration monitoring (large data transfers)
    \item 24/7 security operations center (SOC) escalation for critical alerts
\end{itemize}

% =============================================================================
% 4.8 DATABASE CONNECTION DESIGN
% =============================================================================

\section{Database Connection Design (Neon PostgreSQL)}

The database connection architecture is designed for optimal performance, reliability, and security when connecting the Node.js application layer to Neon PostgreSQL. Connection pooling, proper error handling, and secure configuration are critical components of this design, ensuring efficient resource utilization and consistent performance under varying load.

\subsection{Connection Pooling Architecture}

Connection pooling maintains a cache of database connections that can be reused across multiple client requests, eliminating the overhead of establishing a new connection for each database operation. This is particularly important for serverless and high-concurrency environments where connection establishment costs can significantly impact performance.

\textbf{Pool Configuration:}
\begin{itemize}
    \item \textbf{Min Pool Size}: 5 connections (maintain at least 5 connections even when idle to handle sudden spikes)
    \item \textbf{Max Pool Size}: 20 connections (maximum connections in the pool, limits database load)
    \item \textbf{Connection Timeout}: 10 seconds (time to wait for a connection from pool before error)
    \item \textbf{Idle Timeout}: 30 seconds (close idle connections after this time to free resources)
    \item \textbf{Max Lifetime}: 30 minutes (recycle connections periodically to avoid stale connections)
    \item \textbf{Queue Limit}: 50 (maximum queued requests when pool is exhausted, prevents memory overflow)
\end{itemize}

\textbf{Pool Benefits:}
\begin{itemize}
    \item Reduced latency (no connection setup overhead, saving 20-50ms per request)
    \item Better resource utilization (controlled number of connections, prevents database overload)
    \item Connection limit protection (prevents database from being overwhelmed)
    \item Automatic recovery (replaces failed connections transparently)
    \item Load smoothing (handles traffic spikes gracefully without connection storms)
\end{itemize}

\subsection{Connection Implementation}

The following code implements the connection pool with comprehensive error handling and monitoring:

\begin{verbatim}
// config/database.js - Database connection configuration
require('dotenv').config();
const { Pool } = require('@neondatabase/serverless');

// Create connection pool with optimal configuration
const pool = new Pool({
    connectionString: process.env.DATABASE_URL,
    ssl: {
        require: true,              // Enforce SSL connection
        rejectUnauthorized: true     // Verify server certificate (prevents MITM)
    },
    max: 20,                         // Maximum pool size
    idleTimeoutMillis: 30000,        // Close idle connections after 30s
    connectionTimeoutMillis: 10000,   // Return error after 10s if no connection
    maxUses: 7500,                    // Recycle connection after 7500 uses
});

// Enhanced query function with logging and error handling
const query = async (text, params) => {
    const start = Date.now();
    try {
        const res = await pool.query(text, params);
        const duration = Date.now() - start;
        
        // Log slow queries for optimization (threshold: 100ms)
        if (duration > 100) {
            console.warn('Slow query detected:', { text, duration, rows: res.rowCount });
        }
        
        // Log all queries in development for debugging
        if (process.env.NODE_ENV === 'development') {
            console.log('Executed query:', { text, duration, rows: res.rowCount });
        }
        
        return res;
    } catch (error) {
        // Structured error logging (sanitized, no sensitive data)
        console.error('Database query error:', {
            message: error.message,
            code: error.code,
            detail: error.detail,
            query: text,
            params: params ? params.map(p => typeof p) : [] // Log types only, not values
        });
        
        // Throw normalized error for consistent handling
        throw new DatabaseError(error.message, error.code);
    }
};

// Transaction support with automatic rollback on error
const transaction = async (callback) => {
    const client = await pool.connect();
    try {
        await client.query('BEGIN');
        const result = await callback(client);
        await client.query('COMMIT');
        return result;
    } catch (error) {
        await client.query('ROLLBACK');
        throw error;
    } finally {
        client.release();
    }
};

// Health check function for monitoring
const healthCheck = async () => {
    const start = Date.now();
    try {
        const res = await pool.query('SELECT 1');
        return { 
            status: 'healthy', 
            latency: Date.now() - start,
            poolSize: pool.totalCount,
            idle: pool.idleCount,
            waiting: pool.waitingCount
        };
    } catch (error) {
        return { status: 'unhealthy', error: error.message };
    }
};

// Graceful shutdown for application termination
const closePool = async () => {
    console.log('Closing database connection pool...');
    await pool.end();
    console.log('Database pool closed');
};

// Handle application termination signals
process.on('SIGINT', closePool);
process.on('SIGTERM', closePool);

module.exports = { pool, query, transaction, healthCheck, closePool };
\end{verbatim}

\subsection{Connection Security}

\textbf{SSL/TLS Configuration:}
\begin{itemize}
    \item SSL mode: \texttt{require} (enforces encryption, fails if not available)
    \item Certificate validation: Enabled (prevents MITM attacks, verifies server identity)
    \item Minimum TLS version: TLSv1.2 (disables older, insecure versions)
    \item Cipher suites: Strong ciphers only (AES-GCM, ChaCha20-Poly1305)
    \item Certificate pinning: Optional for additional security (store server certificate hash)
\end{itemize}

\textbf{Credential Management:}
\begin{itemize}
    \item Connection string stored in environment variables (\texttt{DATABASE\_URL})
    \item Never committed to version control (.gitignore enforced)
    \item Rotated quarterly or after any security incident (automated)
    \item Separate credentials for read-only replicas (if used for reporting)
    \item IAM roles for cloud environments (AWS RDS IAM authentication, no static passwords)
\end{itemize}

\subsection{Error Handling and Retry Logic}

\textbf{Connection Errors:}
\begin{itemize}
    \item Transient errors (timeouts, temporary failures): Automatic retry with exponential backoff (100ms, 500ms, 1s, 5s)
    \item Permanent errors (authentication failure, invalid database): Log and alert, no retry (requires human intervention)
    \item Connection exhaustion: Queue requests with timeout, return 503 Service Unavailable with retry-after header
\end{itemize}

\textbf{Query Errors:}
\begin{itemize}
    \item Syntax errors: Return 400 Bad Request (client error, needs correction)
    \item Constraint violations (unique, foreign key): Map to appropriate HTTP status codes (409 Conflict)
    \item Deadlocks: Automatic retry (up to 3 times) with random delay
    \item Serialization failures: Retry transaction with exponential backoff
\end{itemize}

\subsection{Monitoring and Alerting}

\textbf{Metrics Collected:}
\begin{itemize}
    \item Pool size (current, active, idle, waiting) - for capacity planning
    \item Query latency (average, p95, p99) - for performance monitoring
    \item Error rates by error type (syntax, constraint, connection) - for reliability
    \item Connection acquisition wait time - for pool sizing
    \item Transaction throughput (per second/minute) - for load analysis
    \item Connection age and reuse count - for connection health
\end{itemize}

\textbf{Alert Thresholds:}
\begin{itemize}
    \item Pool exhaustion (waiting queries > 10) → Alert (scale needed)
    \item High latency (p95 > 500ms for 5 minutes) → Alert (query optimization needed)
    \item Error rate > 5% over 5 minutes → Alert (investigate)
    \item Connection acquisition timeout > 1s → Warning (pool saturation)
    \item Database disconnection → Critical alert (immediate action)
\end{itemize}

% =============================================================================
% 4.9 THIRD-PARTY INTEGRATIONS
% =============================================================================

\section{Third-party Integrations}

The BondHealth platform integrates with multiple external services to extend functionality, enhance user experience, and leverage specialized capabilities while maintaining focus on core healthcare features. Each integration is carefully selected for reliability, security, and compliance with healthcare regulations, with appropriate fallbacks and error handling.

\subsection{Integration Architecture Principles}

\begin{itemize}
    \item \textbf{Abstraction}: Integration logic encapsulated behind service interfaces, allowing easy replacement of providers
    \item \textbf{Resilience}: Circuit breakers, retries, and fallbacks for all external calls (Hystrix pattern)
    \item \textbf{Security}: API keys stored securely in environment variables or vault, never in code
    \item \textbf{Monitoring}: All external calls tracked for performance, errors, and availability
    \item \textbf{Compliance}: Business Associate Agreements (BAAs) for healthcare data with all vendors
    \item \textbf{Cost Optimization}: Request caching and batching where appropriate to minimize API calls
    \item \textbf{Versioning}: API version pinning to prevent breaking changes
\end{itemize}

\begin{table}[h]
    \centering
    \caption{Third-party Integrations - Comprehensive list of all external services integrated with the BondHealth platform, including purpose, implementation details, and compliance considerations. Each integration is chosen for reliability and healthcare compliance.}
    \label{tab:thirdparty}
    \begin{tabular}{|p{3.5cm}|p{3.5cm}|p{7cm}|}
        \hline
        \textbf{Integration} & \textbf{Purpose} & \textbf{Implementation Details} \\
        \hline
        Stripe/Razorpay & Payment processing & RESTful API integration with webhooks for asynchronous notifications. PCI DSS Level 1 compliant. Supports multiple payment methods (cards, wallets, bank transfers, UPI). Handles subscriptions, one-time payments, and refunds. Webhook endpoints for payment success/failure, disputes, and refunds. Idempotency keys prevent duplicate charges. \\
        \hline
        Twilio/MessageBird & SMS notifications & API integration for sending transactional SMS. Templates for appointment reminders, lab report alerts, and critical notifications. Two-way SMS for patient responses (e.g., confirm appointment). Delivery status tracking with webhooks. Opt-out management (STOP keyword). Compliance with TCPA and local regulations. \\
        \hline
        SendGrid/NodeMailer & Email notifications & Email delivery with templates hosted on SendGrid (dynamic templates). Transactional emails (appointment confirmations, reports) and marketing emails (with consent). Open and click tracking for analytics. DKIM/SPF configuration for deliverability. Bounce handling and suppression lists. Webhooks for delivery events. \\
        \hline
        AWS S3/Azure Blob & File storage & Secure cloud storage for medical reports, images, and documents. Server-side encryption (AES-256). Lifecycle policies for archival (Glacier after 90 days) and deletion (after 7 years). Signed URLs for secure access (5-minute expiry). CORS configuration for direct browser uploads. Versioning enabled for critical documents. \\
        \hline
        Socket.io & Real-time chat & WebSocket library for bidirectional communication. Redis adapter for multi-server scalability (pub/sub). Authentication middleware with JWT. Room-based messaging for privacy. Fallback to long-polling for incompatible clients. Connection state recovery with acknowledgment. \\
        \hline
        Google Calendar API & Calendar sync & OAuth 2.0 integration for adding appointments to patient calendars. Create/update/delete events with reminders. Free/busy checking for availability (optional). Push notifications for changes via webhooks. Supports multiple calendar providers (Google, Apple, Outlook). \\
        \hline
        Firebase Cloud Messaging & Push notifications & Cross-platform push notifications for mobile apps (iOS, Android). Topic-based subscriptions for broadcast messages. Notification payloads with deep linking (open specific screen). Analytics for delivery and open rates. Support for data messages (silent sync). \\
        \hline
        Google Maps API & Location services & Hospital location display with interactive maps. Distance calculation for nearby searches (geolocation). Geocoding for address validation (standardize addresses). Places autocomplete for address entry (reduce errors). \\
        \hline
        Twilio Video/Vonage & Telehealth & Video consultation integration with WebRTC. Recording capabilities (with patient consent). Waiting rooms with notification. Screen sharing for document review. Chat during calls (side channel). Bandwidth adaptation for poor connections. \\
        \hline
        Auth0 (optional) & Identity management & Social login integration (Google, Facebook, Apple). Multi-factor authentication (SMS, authenticator app). Single Sign-On (SSO) for enterprise (SAML). User directory sync with HR systems. \\
        \hline
        Slack/Discord & Internal notifications & Admin alerts for critical events (system errors, urgent reports). Integration with monitoring tools (Prometheus alerts). Incident response notifications with channels. Webhook integration for custom alerts. \\
        \hline
        DataDog/New Relic & Monitoring & Application performance monitoring (APM). Error tracking with stack traces. Custom metrics for business KPIs (appointments, revenue). Dashboard and alerting with Slack integration. Distributed tracing for microservices. \\
        \hline
        Sentry & Error tracking & Real-time error tracking with source maps. User context for debugging (role, ID, but no PII). Issue grouping and tracking with severity. Release tracking for regression detection. \\
        \hline
        Cloudflare & CDN and Security & Content delivery network for static assets (images, CSS, JS). DDoS protection (L3/L7). Web Application Firewall (WAF) with OWASP rules. SSL/TLS termination with edge certificates. \\
        \hline
    \end{tabular}
\end{table}

\subsection{Payment Gateway Integration Details}

\textbf{Stripe Integration:}
\begin{itemize}
    \item \textbf{API Version}: 2023-10-16 (pinned to avoid breaking changes)
    \item \textbf{Authentication}: Secret key stored in environment variables (rotated quarterly)
    \item \textbf{Payment Methods}: Cards, Apple Pay, Google Pay, bank transfers (ACH), Buy Now Pay Later
    \item \textbf{Webhooks}: Payment intent succeeded/failed, refund updated, dispute created, payout completed
    \item \textbf{Idempotency}: Idempotency keys prevent duplicate charges (retry-safe)
    \item \textbf{Receipts}: Automatic email receipts from Stripe (optional, can override)
    \item \textbf{Connect}: For marketplace scenarios (future expansion to multi-hospital)
\end{itemize}

\textbf{Razorpay Integration (India-specific):}
\begin{itemize}
    \item \textbf{API Version}: 1.0
    \item \textbf{Payment Methods}: Cards, UPI, NetBanking, Wallets, EMI, BNPL
    \item \textbf{Webhooks}: Payment authorized/captured/refunded, settlement processed
    \item \textbf{Invoices}: Generate GST-compliant invoices with tax breakdown
    \item \textbf{Settlement}: Automatic settlement to bank accounts (daily/weekly)
\end{itemize}

\subsection{File Storage Integration Details}

\textbf{AWS S3 Configuration:}
\begin{itemize}
    \item \textbf{Bucket Structure}: \texttt{/bucket-name/tenant-id/document-type/year/month/uuid-filename} for organization
    \item \textbf{Encryption}: Server-side encryption with AWS KMS (SSE-KMS) using customer-managed keys
    \item \textbf{Access Control}: Presigned URLs for upload/download (5-minute expiry, single-use)
    \item \textbf{Lifecycle Rules}: Move to Glacier after 90 days (cost optimization), delete after 7 years (compliance)
    \item \textbf{Versioning}: Enabled for audit trail and recovery (30-day version retention)
    \item \textbf{Cross-region replication}: Optional for disaster recovery (secondary region)
\end{itemize}

\textbf{Azure Blob Storage Configuration:}
\begin{itemize}
    \item \textbf{Container}: Private container with SAS tokens for access (time-limited)
    \item \textbf{Access Tier}: Hot for active documents (0-90 days), Cool for archives (90-365 days)
    \item \textbf{Encryption}: Azure Storage Service Encryption with customer-managed keys
    \item \textbf{Soft Delete}: 30-day retention for deleted files (recovery option)
\end{itemize}

\subsection{Notification Service Integration Details}

\textbf{Twilio SMS Integration:}
\begin{itemize}
    \item \textbf{Phone Numbers}: Toll-free numbers for US/CAN, local numbers for other regions
    \item \textbf{Messaging Service}: Pool of numbers for high availability and load balancing
    \item \textbf{Templates}: Pre-approved templates for compliance (A2P 10DLC)
    \item \textbf{Status Callbacks}: Webhooks for delivery status (delivered, failed, undelivered)
    \item \textbf{Opt-out Management}: Automatic handling of STOP/HELP keywords
\end{itemize}

% =============================================================================
% 4.10 PERFORMANCE OPTIMIZATION
% =============================================================================

\section{Performance Optimization}

The BondHealth platform implements multiple optimization strategies to ensure responsive user experiences even under high load, with sub-second response times for critical operations. These optimizations span database, application, and infrastructure layers, ensuring scalability as user base grows.

\textbf{Database Optimization:}
\begin{itemize}
    \item Database indexing on frequently queried columns such as email addresses, user roles, appointment dates, and foreign key relationships significantly reduces query execution times (10-100x improvement)
    \item Connection pooling minimizes database connection overhead and manages concurrent access efficiently (reduces connection establishment from 50ms to near zero)
    \item Query optimization through EXPLAIN analysis and index tuning (regular review)
    \item Read replicas for reporting queries to offload primary database (separate read-only connection)
    \item Partitioning for large tables (appointments by date, logs by month) to improve query performance
\end{itemize}

\textbf{Caching Strategy:}
\begin{itemize}
    \item Redis caching for frequently accessed data such as doctor lists, specializations, and hospital information (5-minute TTL)
    \item Session data stored in Redis for fast access (sub-millisecond)
    \item Response caching for public endpoints (doctor directory, hospital info) with CDN
    \item Database query caching for expensive but static queries (reference data)
    \item Object caching for user profiles after authentication (reduce database hits)
\end{itemize}

\textbf{API Optimization:}
\begin{itemize}
    \item Pagination for all list endpoints including appointments, patients, and reports, ensuring response sizes remain manageable (default 20 items per page)
    \item Field selection (sparse fields) to allow clients to request only needed data
    \item Compression (gzip/brotli) for API responses reduces bandwidth by 70-80%
    \item HTTP/2 multiplexing for parallel requests
    \item Batch endpoints for mobile clients to reduce round trips
\end{itemize}

\textbf{Frontend Optimization:}
\begin{itemize}
    \item Lazy loading for images and file attachments improves initial page load times (images load as user scrolls)
    \item Code splitting for JavaScript bundles (load only what's needed per page)
    \item Asset minification and bundling (Webpack/Rollup)
    \item CDN for static assets (images, CSS, JS) with edge caching (Cloudflare)
    \item Service workers for offline capability and caching (PWA)
\end{itemize}

\textbf{Infrastructure Scaling:}
\begin{itemize}
    \item Horizontal scaling through stateless application design, allowing multiple Node.js instances to handle increased load (auto-scaling based on CPU/memory)
    \item Load balancing with NGINX/HAProxy across instances (round-robin with health checks)
    \item Auto-scaling groups based on metrics (CPU > 70\%, request queue length)
    \item Database read replicas can distribute query load (up to 5 replicas)
    \item CDN for static assets reduces origin load by 90\%
\end{itemize}

\textbf{Monitoring and Alerting:}
\begin{itemize}
    \item Real-time performance monitoring with DataDog/New Relic (dashboards for all metrics)
    \item Alerting on performance degradation (p95 latency > 500ms)
    \item Log aggregation for slow query analysis (ELK stack with Kibana dashboards)
    \item Synthetic monitoring for critical user journeys (booking, login) from multiple locations
    \item Business metrics tracking (appointments per minute, active users) for capacity planning
\end{itemize}