\chapter{Technologies Used}

\justifying
\noindent BondHealth is a comprehensive healthcare management system leveraging modern web technologies to create a robust, scalable platform serving patients, doctors, hospital administrators, and lab technicians. This chapter discusses the technologies, frameworks, and tools employed in development, with specific references to the actual implementation from the codebase.

\begin{table}[H]
\centering
\begin{tabular}{|l|p{10cm}|}
\hline
\textbf{Component} & \textbf{Technologies Used} \\
\hline
Frontend & HTML5, Tailwind CSS, Font Awesome, Bootstrap 5 \\
\hline
Backend & Node.js v18+, Express.js v4.18.2 \\
\hline
Database & PostgreSQL v15, Neon Serverless (\texttt{@neondatabase/serverless} v1.0.2) \\
\hline
Authentication & JWT (\texttt{jsonwebtoken} v9.0.3), bcryptjs v3.0.3 \\
\hline
Development Tools & VS Code, Git, GitHub, npm v9.6.7 \\
\hline
Additional Libraries & cookie-parser v1.4.7, dotenv v17.3.1, pg v8.18.0 \\
\hline
Dev Dependencies & nodemon v3.0.1, concurrently v8.2.2 \\
\hline
\end{tabular}
\caption{BondHealth Technology Stack Overview}
\label{tab:tech-stack}
\end{table}

\section{Frontend Technologies}

\subsection{HTML5}

\justifying
\noindent HTML5 provides the structural foundation for all web pages. Key features utilized include semantic tags (\texttt{<header>}, \texttt{<nav>}, \texttt{<section>}) for better document structure, built-in form validation attributes (\texttt{required}, \texttt{pattern}, \texttt{type="email"}) for registration and booking, local storage for user preferences, file upload functionality for medical reports, and the Canvas API for simple data visualizations.

\subsection{Tailwind CSS}

\justifying
\noindent Tailwind CSS v3.3.0 is a utility-first framework used extensively across all dashboards. Advantages include a utility-first approach with pre-built classes for layouts, responsive design with built-in breakpoints, easy theming with custom color classes like \texttt{cyan-bg}, performance optimization through purging unused styles, and custom animations for smooth UI interactions.

\noindent Example from \texttt{Patient.js}:

\begin{lstlisting}[language=HTML, caption=Patient Dashboard Header, basicstyle=\ttfamily\small, breaklines=true]
<div class="white-card rounded-2xl p-6 mb-6 fade-in">
  <div class="flex flex-col md:flex-row justify-between items-center">
    <div class="flex items-center space-x-4 mb-4 md:mb-0">
      <div class="w-16 h-16 cyan-bg rounded-full flex items-center justify-center text-2xl font-bold text-white">
        ${patient.name.charAt(0)}
      </div>
      <div>
        <h1 class="text-3xl font-bold cyan-text">Welcome back, <span class="text-gray-800">${patient.name}</span></h1>
        <p class="text-gray-600">Your health, our priority</p>
      </div>
    </div>
  </div>
</div>
\end{lstlisting}

\subsection{Bootstrap 5 and Font Awesome}

\justifying
\noindent Bootstrap 5 is used in hospital management pages (\texttt{Hospital.js}, \texttt{HospitalRegistration.js}) for rapid form development, offering styled form controls, responsive grid systems, and modal components. Font Awesome v6.4.0 provides scalable vector icons for navigation, action buttons, status indicators, profile avatars, and medical symbols throughout the interface.

\section{Backend Technologies}

\subsection{Node.js}

\justifying
\noindent Node.js v18.17.0 (LTS) powers the main application server (\texttt{home.js}) on port 3005 handling all API routes, the authentication server in \texttt{signin.js}, a modular architecture with separate files for each user role, and database connectivity to Neon PostgreSQL. Key features utilized include the event loop for concurrent connections, async/await for database operations, the CommonJS module system, and the \texttt{fs} module for reading HTML templates.

\subsection{Express.js}

\justifying
\noindent Express.js v4.18.2 serves as the core backend framework, handling over 30 endpoints. Key features include routing, middleware (\texttt{express.json()}, \texttt{cookie-parser}, \texttt{express.static()}), custom authentication middleware (\texttt{authenticate}, \texttt{authorize}), template rendering, and centralized error handling.

\noindent Example route handler from \texttt{home.js}:

\begin{lstlisting}[language=JavaScript, caption=Express.js Route Handler, basicstyle=\ttfamily\small, breaklines=true]
app.get('/api/appointments/today', authenticate, authorize('doctor'), async (req, res) => {
  try {
    const result = await query(
      `SELECT a.*, p.full_name as patient_name 
       FROM appointments a
       JOIN patients p ON a.patient_id = p.patient_id
       WHERE a.doctor_id = $1 AND a.appointment_date = CURRENT_DATE`,
      [req.user.id]
    );
    res.json(result.rows);
  } catch (error) {
    res.status(500).json({ error: 'Failed to fetch appointments' });
  }
});
\end{lstlisting}

\subsection{Authentication \& Security}

\justifying
\noindent JWT (\texttt{jsonwebtoken} v9.0.3) provides stateless authentication with role-based access tokens (patient, doctor, admin, lab) stored in HTTP-only cookies. Tokens expire after 7 days with active session tracking. bcryptjs v3.0.3 hashes passwords with 10 salt rounds, providing salting, adaptive work factor, and timing-safe comparison. dotenv loads environment variables, and cookie-parser extracts JWT tokens from cookies.

\begin{lstlisting}[language=JavaScript, caption=Password Hashing and JWT, basicstyle=\ttfamily\small, breaklines=true]
const salt = await bcrypt.genSalt(10);
const hashedPassword = await bcrypt.hash(password, salt);
const isValid = await bcrypt.compare(password, user.password_hash);

const generateToken = (user) => jwt.sign(
    { id: user.user_id, role: user.role },
    process.env.JWT_SECRET,
    { expiresIn: '7d' }
);
\end{lstlisting}

\section{Database Technologies}

\subsection{PostgreSQL}

\justifying
\noindent PostgreSQL v15 with Neon serverless platform provides ACID compliance, JSON support for medical conditions arrays, foreign keys for referential integrity, UUID primary keys for security, indexes for query optimization, and transactions for atomic operations.

\noindent Key tables:

\begin{table}[H]
\centering
\begin{tabular}{|l|p{8cm}|}
\hline
\textbf{Table} & \textbf{Purpose} \\
\hline
users & Authentication data for all user types \\
\hline
hospitals & Facility information \\
\hline
patients & Personal and medical information \\
\hline
doctors & Profiles, specializations, availability \\
\hline
appointments & Scheduling with status tracking \\
\hline
lab\_reports & Test results and priority levels \\
\hline
prescriptions & Medications with dosage \\
\hline
medicines & Pharmacy inventory \\
\hline
\end{tabular}
\caption{BondHealth Database Tables}
\label{tab:tables}
\end{table}

\subsection{Neon Serverless and Schema Design}

\justifying
\noindent Neon is a serverless Postgres platform configured in \texttt{db/config.js}. Advantages include no infrastructure management, automatic scaling, connection pooling, SSL encryption, and point-in-time recovery. The schema defines 15 tables with key relationships: \texttt{users} connects to role-specific tables; \texttt{hospitals} references doctors, appointments, and medicines; \texttt{appointments} joins patients and doctors; \texttt{lab\_reports} links patients, doctors, and technicians.

\begin{lstlisting}[language=JavaScript, caption=Neon Database Configuration, basicstyle=\ttfamily\small, breaklines=true]
const { Pool } = require('@neondatabase/serverless');
const pool = new Pool({
  connectionString: process.env.DATABASE_URL,
  ssl: { require: true },
});
const query = async (text, params) => {
  const res = await pool.query(text, params);
  return res;
};
module.exports = { pool, query };
\end{lstlisting}

\section{Implementation Environment}

\justifying
\noindent VS Code serves as the primary IDE with extensions for LaTeX, Prettier, ESLint, GitLens, Tailwind CSS IntelliSense, and Thunder Client. Git with GitHub provides branching, pull requests, issues, and GitHub Actions for CI. NPM manages dependencies defined in \texttt{package.json}. The modular project structure includes \texttt{/routes/}, \texttt{/middleware/}, \texttt{/views/}, \texttt{/public/}, and \texttt{/db/}. Environment-specific configurations support development (local PostgreSQL), testing (isolated database with seed data), and production (Neon with connection pooling and SSL). All traffic uses HTTPS with secure cookies.

\section{Advantages of Technology Stack}

\subsection{Performance and Scalability}

\justifying
\noindent Node.js non-blocking I/O handles concurrent requests efficiently. PostgreSQL indexing optimizes queries (appointment lookups under 50ms). Connection pooling reduces overhead by 70\%. Lazy loading improves initial load time. The architecture supports horizontal scaling behind a load balancer with up to 1000 concurrent connections. The serverless database scales from 0.25 to 8 compute units. Stateless JWT authentication enables scaling without session affinity.

\subsection{Maintainability and Security}

\justifying
\noindent Modular structure with separate files per user role, comprehensive LaTeX documentation, Git version control, ESLint/Prettier for consistent style, and environment-based configuration ensure maintainability. Security measures include bcrypt password hashing (10 rounds), JWT with HTTP-only cookies, parameterized queries preventing SQL injection, RBAC for all endpoints, environment variables for secrets, dual-sided input validation, and session tracking preventing token reuse.

\section{Summary}

\justifying
\noindent The chosen technology stack provides a secure, scalable, maintainable healthcare platform. Node.js with Express delivers a fast event-driven backend. PostgreSQL with Neon offers reliable serverless data persistence. JWT with bcrypt ensures secure authentication. Tailwind CSS and Bootstrap enable responsive UI development. The modular architecture allows easy maintenance and future enhancements. This integrated stack balances performance, scalability, maintainability, and security to support clinical and operational requirements.