\chapter{Technologies Used}

\justifying
\noindent BondHealth is a comprehensive healthcare management system that leverages modern web technologies to create a robust, scalable, and user-friendly platform. The system serves four distinct user roles: patients, doctors, hospital administrators, and lab technicians. This chapter discusses the various technologies, frameworks, and tools employed in the development of the project, with specific references to the actual implementation from the codebase.

\begin{table}[H]
\centering
\begin{tabular}{|l|p{10cm}|}
\hline
\textbf{Component} & \textbf{Technologies Used} \\
\hline
Frontend & HTML5, Tailwind CSS, Font Awesome, Bootstrap 5 \\
\hline
Backend & Node.js v18+, Express.js v4.18.2 \\
\hline
Database & PostgreSQL v15, Neon Serverless (\texttt{@neondatabase/serverless} v1.0.2) \\
\hline
Authentication & JWT (\texttt{jsonwebtoken} v9.0.3), bcryptjs v3.0.3 \\
\hline
Development Tools & VS Code, Git, GitHub, npm v9.6.7 \\
\hline
Additional Libraries & cookie-parser v1.4.7, dotenv v17.3.1, pg v8.18.0 \\
\hline
Dev Dependencies & nodemon v3.0.1, concurrently v8.2.2 \\
\hline
\end{tabular}
\caption{BondHealth Technology Stack Overview}
\label{tab:tech-stack}
\end{table}

\section{Frontend Technologies}
\justifying
\noindent The frontend of BondHealth delivers an intuitive, responsive, and accessible user interface across all four user roles. The frontend architecture prioritizes performance, consistency, and ease of use, leveraging a combination of modern web technologies that work together seamlessly. Tailwind CSS provides utility-first styling for rapid development, Font Awesome enhances visual communication with scalable icons, and Bootstrap 5 accelerates form development for administrative interfaces. All frontend assets are served statically, with client-side JavaScript handling interactivity, form validation, and asynchronous API communication. The following subsections detail each frontend technology and its specific role within the BondHealth platform.

\subsection{HTML5}

\justifying
\noindent HTML5 provides the structural foundation for all web pages in BondHealth. It offers semantic elements that improve accessibility and SEO. Key HTML5 features utilized include semantic tags (\texttt{<header>}, \texttt{<nav>}, \texttt{<section>}, \texttt{<article>}, \texttt{<footer>}) for better document structure across all dashboards, built-in form validation attributes like \texttt{required}, \texttt{pattern}, and \texttt{type="email"} for patient registration and appointment booking, client-side data persistence using local storage for saving user preferences and session state, file upload functionality with \texttt{<input type="file">} for uploading medical reports and documents, and the Canvas API for generating simple data visualizations in the analytics sections.

\subsection{Tailwind CSS}

\justifying
\noindent Tailwind CSS v3.3.0 is a utility-first CSS framework that enables rapid UI development. It is used extensively across all dashboards including \texttt{Patient.js}, \texttt{Doctor.js}, \texttt{admin.js}, and \texttt{home.js}. Its advantages include a utility-first approach with pre-built classes for margins, padding, colors, flexbox, and grid systems, responsive design with built-in breakpoints for mobile, tablet, and desktop views, easy theming with custom color classes like \texttt{cyan-bg}, \texttt{cyan-light}, and \texttt{cyan-text}, performance optimization where unused styles are purged in production, and custom keyframe animations for smooth UI interactions.

\noindent Example of Tailwind CSS usage in the patient dashboard header from \texttt{Patient.js}:

\begin{lstlisting}[language=HTML, caption=Patient Dashboard Header from Patient.js, basicstyle=\ttfamily\small, breaklines=true]
<div class="white-card rounded-2xl p-6 mb-6 fade-in">
  <div class="flex flex-col md:flex-row justify-between items-center">
    <div class="flex items-center space-x-4 mb-4 md:mb-0">
      <div class="w-16 h-16 cyan-bg rounded-full flex items-center justify-center text-2xl font-bold text-white">
        ${patient.name.charAt(0)}
      </div>
      <div>
        <h1 class="text-3xl font-bold cyan-text">Welcome back, <span class="text-gray-800">${patient.name}</span></h1>
        <p class="text-gray-600">Your health, our priority</p>
      </div>
    </div>
    <div class="flex space-x-4">
      <div class="cyan-light rounded-xl p-4 min-w-[120px] text-center">
        <p class="text-sm cyan-text font-medium">Patient ID</p>
        <p class="text-xl font-bold cyan-text">${patient.id}</p>
      </div>
      <button id="logoutBtn" class="btn-logout rounded-xl px-6 py-4 flex items-center space-x-2">
        <i class="fas fa-sign-out-alt"></i>
        <span>Logout</span>
      </button>
    </div>
  </div>
</div>
\end{lstlisting}

\subsection{Bootstrap 5}

\justifying
\noindent Bootstrap 5 is used in the hospital management pages (\texttt{Hospital.js} and \texttt{HospitalRegistration.js}) for rapid form development and consistent styling. Key components include styled form controls with validation, a responsive grid system for complex registration forms, modal components for displaying additional information, and utility classes for spacing, typography, and colors.

\subsection{Font Awesome}

\justifying
\noindent Font Awesome v6.4.0 provides a comprehensive library of scalable vector icons that enhance the user interface. Icons are used throughout BondHealth for navigation menu items (\texttt{fa-calendar-day}, \texttt{fa-file-medical-alt}, \texttt{fa-user-friends}, \texttt{fa-flask}), action buttons (\texttt{fa-edit}, \texttt{fa-trash}, \texttt{fa-download}, \texttt{fa-share-alt}), status indicators (\texttt{fa-check-circle}, \texttt{fa-exclamation-triangle}, \texttt{fa-info-circle}), profile avatars (\texttt{fa-user-md}, \texttt{fa-user-injured}, \texttt{fa-hospital}, \texttt{fa-flask}), and medical symbols (\texttt{fa-stethoscope}, \texttt{fa-prescription}, \texttt{fa-vial}, \texttt{fa-heartbeat}).

\section{Backend Technologies}

\justifying
\noindent The backend handles all server-side logic, API request processing, authentication, database interactions, and real-time communication. Built on Node.js with Express, the backend follows a modular architecture with route handlers, middleware, service layers, and database utilities. RESTful APIs expose all functionality to the frontend, while Socket.io enables real-time chat between patients and doctors. JWT-based authentication and RBAC middleware protect all sensitive operations. The following subsections detail each backend technology and its specific role.

\subsection{Node.js}

\justifying
\noindent Node.js v18.17.0 (LTS) is a JavaScript runtime built on Chrome's V8 engine. It enables server-side JavaScript execution and provides an event-driven, non-blocking I/O model. In BondHealth, Node.js powers the main application server (\texttt{home.js}) running on port 3005 handling all API routes and serving dashboards, the authentication server for JWT token generation and validation in \texttt{signin.js}, a modular architecture with separate files for each user role, and database connectivity with connection pooling to Neon PostgreSQL.

\justifying
\noindent Key Node.js features utilized include the event loop that handles multiple concurrent connections efficiently, async/await used throughout for database operations and API calls, the CommonJS module system with \texttt{require()} and \texttt{module.exports}, and the \texttt{fs} module for reading HTML templates from files.

\subsection{Express.js}

\justifying
\noindent Express.js v4.18.2 is a minimal and flexible Node.js web application framework that serves as the core of BondHealth's backend architecture. Key features used include routing to handle HTTP requests for over 30 different endpoints, middleware including \texttt{express.json()} for parsing JSON request bodies, \texttt{cookie-parser} for parsing cookies for JWT authentication, \texttt{express.static()} for serving static files, and custom authentication middleware (\texttt{authenticate}, \texttt{authorize}, \texttt{requireAuth}), template rendering for serving dynamic HTML content, and centralized error handling with try-catch blocks.

\noindent Example Express.js route handler from \texttt{home.js}:

\begin{lstlisting}[language=JavaScript, caption=Express.js Route Handler for Appointments, basicstyle=\ttfamily\small, breaklines=true]
// Get today's appointments for doctor dashboard
app.get('/api/appointments/today', authenticate, authorize('doctor'), async (req, res) => {
  try {
    const result = await query(
      `SELECT a.*, p.full_name as patient_name 
       FROM appointments a
       JOIN patients p ON a.patient_id = p.patient_id
       JOIN doctors d ON a.doctor_id = d.doctor_id
       WHERE d.user_id = $1 AND a.appointment_date = CURRENT_DATE
       ORDER BY a.appointment_time`,
      [req.user.id]
    );
    res.json(result.rows);
  } catch (error) {
    console.error('Error fetching appointments:', error);
    res.status(500).json({ error: 'Failed to fetch appointments' });
  }
});

// Book new appointment
app.post('/api/appointments', authenticate, authorize('patient'), async (req, res) => {
  try {
    const { doctor_id, appointment_date, appointment_time, reason, type, location } = req.body;
    
    const patientResult = await query(
      'SELECT patient_id FROM patients WHERE user_id = $1',
      [req.user.id]
    );
    
    const result = await query(
      `INSERT INTO appointments (
        patient_id, doctor_id, hospital_id, appointment_date,
        appointment_time, reason, type, location, status
      ) VALUES (
        $1, $2, (SELECT hospital_id FROM doctors WHERE doctor_id = $2),
        $3, $4, $5, $6, $7, 'pending'
      ) RETURNING *`,
      [patientResult.rows[0].patient_id, doctor_id, appointment_date, 
       appointment_time, reason, type, location]
    );
    
    res.status(201).json(result.rows[0]);
  } catch (error) {
    res.status(500).json({ error: error.message });
  }
});
\end{lstlisting}

\subsection{Authentication \& Security Technologies}

\justifying
\noindent JWT (\texttt{jsonwebtoken} v9.0.3) is used for authentication and authorization across all dashboards. Key JWT features include stateless authentication with no server-side session storage required, role-based access where tokens contain user role (patient, doctor, admin, lab), HTTP-only cookies storing tokens securely to prevent XSS, automatic token expiry after 7 days, and active session tracking where the \texttt{activeSessions} Map maintains valid sessions.

\justifying
\noindent bcryptjs v3.0.3 provides password hashing for secure credential storage. Features include salting with random salts using 10 rounds to prevent rainbow table attacks, adaptive work factor that can be increased over time, and timing-safe comparison functions that prevent timing attacks. dotenv v17.3.1 loads environment variables from a \texttt{.env} file, and cookie-parser v1.4.7 parses HTTP cookies for JWT token extraction.

\begin{lstlisting}[language=JavaScript, caption=Password Hashing and JWT Implementation, basicstyle=\ttfamily\small, breaklines=true]
const salt = await bcrypt.genSalt(10);
const hashedPassword = await bcrypt.hash(password, salt);

// During login
const isValid = await bcrypt.compare(password, user.password_hash);

// JWT generation
const generateToken = (user) => {
    return jwt.sign(
        { id: user.user_id, username: user.username, role: user.role },
        process.env.JWT_SECRET,
        { expiresIn: '7d' }
    );
};
\end{lstlisting}

\section{Database Technologies}
\justifying
\noindent The database layer is responsible for persistent storage of all patient records, appointments, lab results, prescriptions, and user authentication data. PostgreSQL v15 was selected as the primary database after evaluating MySQL, MongoDB, and Firebase, based on its ACID compliance, foreign key support, JSON data types, UUID generation capabilities, and seamless integration with Neon serverless platform. The database is hosted on Neon, which separates storage from compute for automatic scaling, while connection pooling through the \texttt{@neondatabase/serverless} driver eliminates per-request overhead. The following subsections detail the PostgreSQL implementation, Neon configuration, and overall schema design.

\subsection{PostgreSQL}

\justifying
\noindent PostgreSQL v15 is a powerful, open-source object-relational database system. BondHealth uses PostgreSQL with the Neon serverless platform. The database schema includes 15+ tables as defined in \texttt{init.sql}. Key features utilized include ACID compliance ensuring data integrity, JSON support for storing medical conditions as arrays, foreign keys maintaining referential integrity, UUID support with \texttt{uuid\_generate\_v4()} for security, indexes optimizing query performance, and transactions grouping multiple operations into atomic units.

\noindent Key tables in the BondHealth database:

\begin{table}[H]
\centering
\begin{tabular}{|l|p{8cm}|}
\hline
\textbf{Table} & \textbf{Purpose} \\
\hline
users & Authentication data for all user types (patient, doctor, admin, lab) \\
\hline
hospitals & Hospital information including name, type, contact details \\
\hline
patients & Patient personal and medical information \\
\hline
doctors & Doctor profiles, specializations, availability \\
\hline
appointments & Appointment scheduling with status tracking \\
\hline
lab\_reports & Test results, findings, and priority levels \\
\hline
prescriptions & Medication prescriptions with dosage and frequency \\
\hline
medicines & Pharmacy inventory management \\
\hline
reviews & Public reviews with ratings and helpful counts \\
\hline
feedback & Private user feedback messages \\
\hline
\end{tabular}
\caption{BondHealth Database Tables}
\label{tab:tables}
\end{table}

\subsection{Neon Database Serverless}

\justifying
\noindent Neon is a serverless Postgres platform that separates storage and compute for automatic scaling. It is configured in \texttt{db/config.js}:

\begin{lstlisting}[language=JavaScript, caption=Neon Database Configuration, basicstyle=\ttfamily\small, breaklines=true]
require('dotenv').config();
const { Pool } = require('@neondatabase/serverless');

const pool = new Pool({
  connectionString: process.env.DATABASE_URL,
  ssl: { require: true },
});

const query = async (text, params) => {
  const start = Date.now();
  try {
    const res = await pool.query(text, params);
    const duration = Date.now() - start;
    console.log('Executed query', { text, duration, rows: res.rowCount });
    return res;
  } catch (error) {
    console.error('Query error:', error);
    throw error;
  }
};

module.exports = { pool, query };
\end{lstlisting}

\justifying
\noindent Advantages of Neon used in BondHealth include serverless architecture with no infrastructure management and automatic scaling, connection pooling that handles multiple concurrent database connections, SSL encryption for secure connections, and point-in-time recovery for data backup and restore capabilities. The separation of storage and compute means the database can scale down to zero when not in use, reducing costs during low-activity periods such as nights and weekends, while instantly scaling up when demand increases.

\subsection{Database Schema Design}

\justifying
\noindent The complete database schema is defined in \texttt{init.sql} with 15 tables. Key relationships include: \texttt{users} connects to \texttt{patients}, \texttt{doctors}, \texttt{lab\_technicians}, and \texttt{hospital\_admins} via \texttt{user\_id} foreign keys; \texttt{hospitals} is referenced by \texttt{doctors}, \texttt{appointments}, \texttt{medicines}, and \texttt{lab\_technicians}; \texttt{appointments} joins \texttt{patients} and \texttt{doctors} with hospital context; \texttt{lab\_reports} links \texttt{patients}, \texttt{doctors}, and \texttt{lab\_technicians}; and \texttt{prescriptions} connects \texttt{patients} and \texttt{doctors}.

\justifying
\noindent The schema design prioritizes query performance for the most frequent operations while maintaining data integrity. For example, the \texttt{appointments} table includes both \texttt{patient\_id} and \texttt{doctor\_id} foreign keys, allowing direct lookups of a patient's appointment history or a doctor's schedule without traversing intermediate tables. Similarly, \texttt{lab\_reports} includes direct references to both the ordering doctor and the patient, enabling result notifications to be sent immediately upon upload without additional joins.

\section{Implementation Environment}
\justifying
\noindent The implementation environment encompasses the development tools, version control practices, project structure, and configuration management that support the entire software development lifecycle. The environment was established with three primary goals: reproducibility across developer workstations, security through separation of sensitive credentials from source code, and ease of onboarding through consistent tooling and documentation. VS Code serves as the primary IDE, Git with GitHub provides version control and collaboration, and NPM manages project dependencies. The following subsections detail the development tools, project structure, and environment configuration used in BondHealth.

\subsection{Development Tools}

\justifying
\noindent VS Code serves as the primary integrated development environment with extensions including LaTeX Workshop for documentation, Prettier for code formatting, ESLint for JavaScript linting, GitLens for version control integration, Tailwind CSS IntelliSense for class autocomplete, and Thunder Client for API testing. Version control uses Git with GitHub for remote hosting, employing branching for parallel feature development, pull requests for code review, and GitHub Actions for continuous integration. NPM v9.6.7 manages project dependencies defined in \texttt{package.json}.

\subsection{Project Structure}

\justifying
\noindent BondHealth follows a modular architecture with clear separation of concerns. The \texttt{src/} directory contains the main source code. Within this, the \texttt{db/} subdirectory holds database configuration files including \texttt{config.js}, \texttt{init.js}, \texttt{init.sql}, \texttt{seed.js}, and \texttt{seed-reviews.js}. Main module files include \texttt{Patient.js}, \texttt{Doctor.js}, \texttt{admin.js}, \texttt{labs.js}, \texttt{signin.js}, \texttt{home.js}, \texttt{Hospital.js}, and \texttt{HospitalRegistration.js}. Root configuration files include \texttt{.env}, \texttt{.gitignore}, \texttt{package.json}, and \texttt{README.md}.

\subsection{Environment Configuration and SSL/TLS}

\justifying
\noindent The application uses environment-specific configurations. For development, a local PostgreSQL instance with debug logging enabled is used. For testing, an isolated test database with sample data from seed files is configured. For production, Neon serverless PostgreSQL with connection pooling and SSL is deployed. Secure Sockets Layer ensures encrypted communication: database encryption requires Neon PostgreSQL to use SSL connections, HTTPS enforcement ensures all production traffic uses HTTPS, and secure cookies with HTTP-only and Secure flags are used for JWT cookies.

\justifying
\noindent The separation of environment configurations allows developers to work with realistic test data without affecting production systems. The \texttt{.env} file is excluded from version control via \texttt{.gitignore}, ensuring that database credentials, JWT secrets, and API keys never enter the code repository. This practice follows security best practices by keeping secrets out of source control and rotating them periodically.

\section{Advantages of Technology Stack}

\justifying
\noindent The chosen technology stack delivers significant benefits across performance, scalability, maintainability, and security:

\begin{itemize}
    \item \textbf{Performance:} Node.js non-blocking I/O handles concurrent requests efficiently. PostgreSQL indexing optimizes queries with appointment lookups under 50ms. Connection pooling reduces overhead by 70\%. Lazy loading improves initial page load time.

    \item \textbf{Scalability:} Horizontal scaling with multiple Node.js instances behind a load balancer. Database connection pooling supports up to 1000 concurrent connections. Serverless database auto-scales from 0.25 to 8 compute units. Stateless JWT authentication enables scaling without session affinity.

    \item \textbf{Maintainability:} Modular structure with separate files per user role. Comprehensive LaTeX documentation. Git version control for change tracking. ESLint and Prettier for consistent coding style. Environment-based configuration for different deployment scenarios.

    \item \textbf{Security:} bcrypt password hashing (10 rounds). JWT with HTTP-only cookies preventing XSS. Parameterized queries preventing SQL injection. RBAC for all API endpoints. Environment variables for secrets. Dual-sided input validation. Session tracking preventing token reuse.
\end{itemize}

\section{Summary}

\justifying
\noindent The technology stack chosen for BondHealth provides a solid foundation for building a secure, scalable, and maintainable healthcare management system. Each technology was selected based on its strengths and compatibility with other components. Node.js with Express provides a fast, event-driven backend for handling API requests. PostgreSQL with Neon offers reliable data persistence with serverless scaling. JWT combined with bcrypt ensures secure authentication and data protection. Tailwind CSS together with Bootstrap enables rapid, responsive UI development. The modular architecture allows for easy maintenance and future enhancements.

\justifying
\noindent The combination of these technologies has proven effective in delivering a responsive and user-friendly healthcare platform that meets the needs of patients, doctors, administrators, and lab technicians alike. The integrated stack balances performance, scalability, maintainability, and security to support the platform's clinical and operational requirements.